### 2017-08-20 latex/article.tex
added basic files for article

\documentclass[12pt,fleqn]{article}
\usepackage{xiiiemc}
\usepackage{natbib}
\usepackage{fancyhdr}
\usepackage{fancyvrb}
\usepackage{color}
\usepackage{wallpaper} 
\usepackage{titlesec}   %% Define space between paragraph e section
\usepackage{float} 	%% Use to fix Figure or Table: ex: \begin{table}[H]
\usepackage[usenames,dvipsnames]{xcolor}
\usepackage{hyperref} 	%% Use to fix Figure or Table: ex: \begin{table}[H]
\hypersetup{
	%pagebackref=true,
	pdfcreator={LaTeX with abnTeX2},
	pdfkeywords={abnt}{latex}{abntex}{USPSC}{trabalho acadêmico}, 
	colorlinks=true,       		% false: boxed links; true: colored links
	linkcolor=blue,          	% color of internal links
	citecolor=blue,        		% color of links to bibliography
	filecolor=magenta,      		% color of file links
	urlcolor=blue,
	allbordercolors=black,
	bookmarksdepth=4
}
\usepackage{tocloft}
\titleformat{\section}
  {\normalfont\bfseries}{\thesection.}{0.5em}{}
\renewcommand\cftsecaftersnum{.} 
\renewcommand\thesection{\arabic{section}}
\renewcommand\thesubsection{\thesection.\arabic{subsection}}
%%%%Don't edit this block. It reduces the spacing between the lines of the references
\let\OLDthebibliography\thebibliography
\renewcommand\thebibliography[1]{\OLDthebibliography{#1} \setlength{\parskip}{0pt}\setlength{\itemsep}{0pt plus 0.3ex}}

%%-----------------------------------------------EDIT-----------------------------------------------
\title{NECESSARY ENHANCEMENTS OF LINKED DATA EXPRESSIVENESS FOR ONTOLOGIES}

%%-----------------------------------------------EDIT----------------------------------------------
\author
    {\rm \begin{tabular}{l} 
    \textbf{Renato Fabbri}$^{1}$ - {\textnormal renato.fabbri@gmail.com}\\%
%    \textbf{Marília M. Pisani}$^{2}$ - {\textnormal marilia.m.pisani@gmail.com}\\
    {\fontsize{11}{0}\selectfont $^{1}$University of São Paulo, Institute of Mathematical and Computer Sciences - São Carlos, SP, Brazil}\vspace*{-0.05cm} \\
%    {\fontsize{11}{0}\selectfont $^{2}$Federal University of ABC, Centre for Natural Sciences and Humanities - São Paulo, SP, Brazil}\vspace*{-0.05cm}\\
  \end{tabular}}
%%----------------------------------------------------------------------------------------------

\fancypagestyle{firspagetstyle}
{
	\lhead{}
	\fancyhead[C]{%
		\includegraphics[width=0.9\linewidth]{logo}\\%
		{\scriptsize \fontfamily{phv}\fontseries{b}\selectfont \color[rgb]{0.45,0.45,0.45}
		16 a 19 de Outubro de 2017\\
		Instituto Politécnico - Universidade do Estado de Rio de Janeiro\\
		Nova Friburgo - RJ\\
	    }
	}
	\renewcommand{\headrulewidth}{0.0pt}
	\fancyfoot[C]{\footnotesize \parbox{15cm} {\centering  \fontsize{7.5}{0}\selectfont \it Anais do XX ENMC – Encontro Nacional de Modelagem Computacional e VIII ECTM – Encontro de Ciências e Tecnologia de Materiais,  Nova Friburgo, RJ – 16 a 19 Outubro 2017}} % \ttfamil
	\rhead{}
}

\begin{document}
\maketitle

\thispagestyle{firspagetstyle}

\fancyhead[L]{\footnotesize{\fontsize{7.5}{0}\selectfont \it XX ENMC e VIII ECTM\\
	16 a 19 de Outubro de 2017\\
	Instituto Politécnico Universidade do Estado do Rio de Janeiro – Nova Friburgo - RJ\\}}
\renewcommand{\headrulewidth}{0.0pt}
\fancyfoot[C]{\footnotesize \parbox{15cm} {\centering  \fontsize{7.5}{0}\selectfont \it Anais do XX ENMC – Encontro Nacional de Modelagem Computacional e VIII ECTM – Encontro de Ciências e Tecnologia de Materiais,  Nova Friburgo, RJ – 16 a 19 Outubro 2017}} % \ttfamil
\rhead{}

\begin{abstract}
	The semantic web has received many contributions of researchers as ontologies which,
	in this context, i.e. within RDF linked data, are formalized conceptualization that might use
	different protocols, such as RDFS, OWL DL and OWL FULL.
	In this article we describe new expressive techniques which were found necessary after
	elaborating dozens of OWL ontologies for the scientific academy, the State and the civil society.
	They consist in: 1) stating possible uses a property might have without incurring into axioms or restrictions;
	2) assigning a level of priority for an element (class, property, triple); 3) correct depiction in diagrams
	of relations between classes, between individuals which are imperative, and between individuals which are dispensable;
	4) association between OWL classes and SKOS concepts.
	We propose specific rules to accomplish these enhancements and exemplify both its use
	and the difficulties that arise because these techniques are currently not established as basic to the ontology designer.  
\end{abstract}

\keywords{\em{ Semantic web, Linked data, Ontology, RDFS, OWL}}

\pagestyle{fancy}

\section{INTRODUCTION}
The semantic web is constituted by data which are linked in the same way
as web pages are: through HTTP and URLs.
W3C recommendations are the main sources of protocols and best practices of the field.
In practice, the terms 'linked data' and 'semantic web' are most often used interchangeably.
A distinction might arise in some contexts where on need to refer to the linked data or
the semantic web created by all or some portion of linked data, but, as a knowledge and technological
field, the terms are equivalent.

The semantic web is built using the Resource Description Framework (RDF).
The RDF data model is based on making statements in the form of triples
(``subject-predicate-object'') and using Unique Resource Identifier for
objects and concepts.
It is also part of the framework to use URIs that are URLs whenever possible,
to enable the data linkage.
Accordingly, one can write:
\begin{Verbatim}[fontsize=\footnotesize]
<http://example.org/people/mary>
	<http://example.org/properties/name> "Mary Shastacian" .
<http://example.org/people/mary>
	<http://example.org/properties/age> "57" .
<http://example.org/people/mary> 
	<http://example.org/properties/likes> 
                <http://example.org/concepts/reading> .
\end{Verbatim}
\noindent to express that there is person called Mary Shastacian which is 57 years old and likes reading.
There are many formats to write/serialize RDF data.
The example above is written in Turtle and we will use this format throughout this document.

One of the most important notions of the semantic web is that of an ontology.
An ontology, in this context, is a formalized conceptualizations, comprised by
concepts and relations between the concepts and between the relations themselves.
In current semantic web, most simple ontologies are written using the RDFS protocol,
by which one can specify, among other things:
\begin{itemize}
	\item concepts;
	\item properties, which are concepts that are used as predicates and thus relate concepts;
	\item special relations between concepts that state that one concept is mode general than the other\footnote{This
		kind of relation is called hypernymy. Examples: mammal is a hypernym of monkey, drink is a hypernym of beer.};
	\item the subjects and objects that can occurs in a triple where a specific property is a predicate.
\end{itemize}

One can also write an ontology by using OWL,
in which all the expressive capabilities of RDFS are
available, but one can also, among other things:
\begin{itemize}
	\item state ``property axioms'', i.e. specify if a property is e.g. reflexive or transitive;
	\item state ``class restrictions'', i.e. specify if a class instance e.g. necessarily hold a relation to another class instance or data.
\end{itemize}
OWL has a richer vocabulary than RDFS and is (way more) complex.
This complexity is a drawback together with the greater computational cost
for performing inference.
The advantage is the greater power to represent conceptualizations.

The semantic web practices evolves by constant practice and recommendations written
by many specialists usually after years of research and discussion.
Therefore, this document is written in a very humble manner,
without the pretense of yielding new recommendations abruptly,
but as a way to report issues and candidate workarounds/solutions
that emerged by years of intense practice in designing ontologies~\citep{pnud5,fabbri1,losd,ops}.

Section~\ref{sec:prob} is dedicated to exposing some of the problems
in current RDFS and OWL ontologies, with emphasis on the ones
to which this document holds proposals to resolve or lessen.
In Section~\ref{sec:prop} such proposals are described and discussed.
Final remarks are articulated in Section~\ref{sec:con}.

\section{THE PROBLEMS}\label{sec:prob}
The semantic web technology standards are very appreciated by the researchers, developers and users~\citep{semApr}.
Even so, there are many issues regarding current state of the art.
For example, if an individual of a class is found also to be a class,
it implies OWL Full which is a setting presently regarded as not computationally feasible.
This document is dedicated to proposals that resolve (or alleviate)
the problems described in the following subsections.

\subsection{No (standard) way to express potential relations}\label{sec:rel}
There is no recommendation about how to bind to a property a \emph{possible}
subject or object.
Using \texttt{rdfs:doman} and \texttt{rdfs:range} has the consequence that
\textbf{any} subject or object will have be a member of the class specified by
range or domain.
There are, however, many situations where this is not the case.
For example, one might find or create a property \texttt{example:name}.
Suppose it is intended to be used to express names of animals,
say because the person created the property while designing an ontology about animals.
If that is formalized in the property by \texttt{rdfs:domain},
it implies that whenever the \texttt{example:name} is used, the subject
will be inferred (e.g. by an automated reasoner) to be an animal.
Now suppose someone is describing a tool in RDF and wants to write its name.
If he/she uses the \texttt{example:name} property, it will be erroneously
expressing that the tool is an animal.
Such a person is required to create another property \texttt{example2:name}
to write the names.
What is worse: if the person wants to create an ontology about tools,
and wants to express that tools are likely, but not required, to have names,
the property \texttt{exampleX:name} with domain tool will have the same problem
as above, but now with tools instead of animals.
If \textbf{all} the tools have, without exception, a name, then one can express
it as an OWL existential class restriction.
But, again, that is often not the case. Example: algorithms are likely to have names,
but not all of them have names.
One workaround to this problem is to describe these potential relations as notes to the
classes and properties, but it does not favor the machine readable design of the semantic web.

\subsection{All classes and properties are equally fundamental}\label{sec:fun}
The relevance of this issue can be grasped by
tacking the expression of conceptualizations for humans.
Say someone wrote an ontology about machine learning and that it is very complete
in the sense that it contains many concepts, all those that the author could find.
It is straightforward to acknowledge that it will be very difficult for a newcomer
to make sense of everything.
Also, in many cases it is difficult for anyone to make sense of an ontology when it
comprises many concepts and relations.
Thus, it is very useful to have a standard way to express that some set of
concepts or triples are more fundamental then others in a layered fashion,
but there isn't a standard way.

\subsection{Distinction in diagrams of relations between classes and individuals and between imperative and optional relations}\label{sec:dia}
This is not a problem found in formalizing the ontology,
but a problem that arises when depicting the ontology as a diagram.
For example, the (class of) dinosaurs were succeeded by the (class of) mammals,
this is a relation between classes.
A specific animal might have siblings, this is a relation between individuals.
Also, a relation might be imperative for a class (i.e. it might be expressed as a class restriction)
or optional.
These are all poorly distinguished in diagrams that are meant to represent the ontologies.

\subsection{No obvious way to associate an OWL ontology and a SKOS vocabulary}
In a SKOS vocabulary, a concept is an individual of the class \texttt{skos:concept}.
In an OWL ontolgy, a concept is a class.
As stated in the beginning of this section, 
if an individual of a class is also a class,
one has an OWL Full context, which is
currently considered to require inference that is not computable.
This implies that one has to choose between making
an OWL ontology or a SKOS vocabulary or find a way to
make them compatible.
This might be the most complex issue tackled in this document.

\section{ENHANCEMENT PROPOSALS}\label{sec:prop}
As the problems stated in the previous section are distinct,
the proposed enhancements are described case-by-case.

\subsection{Potential domain, range, predicate and post predicate}
To address the problem described in Section~\ref{sec:rel},
we propose creating a new set of properties:
\begin{itemize}
	\item potential/nonrestrictive domain and range of a property might be expressed using\\
		\texttt{example:potentialDomain} and \texttt{example:potentialRange}.
	\item potential predicates when an instance of a class is a subject or an object might
		be expressed using\\
		\texttt{example:potentialPredicate} and \texttt{example:potentialPostPredicate}.
\end{itemize}

The \texttt{example:} in these triples should be replaced by a true namespace prefix, such
as \texttt{http://purl.org/socialparticipation/rdfe/}.
This was not done until the moment because we chose to wait a little longer,
present this article and distribute it online to find out if there is already some
solution currently in use and if there is some trustworthy maintainer that is willing
to keep these properties online in a more suitable namespace.

\subsection{Ascribing numeric values that reflect how fundamental is something}\label{sec:props}
To address the problem described in Section~\ref{sec:fun},
every property and class might me a subject to a triple:
\begin{Verbatim}[fontsize=\footnotesize]
	example:something example:fundamental 4 .
\end{Verbatim}
\noindent with ``4'' replaced by any other numerical value.
At first, it seems reasonable to assume that these values are
valid within the ontology and relative to each other.
Accordingly, one might ascribe \texttt{fundamental 100}
some kernel of meaning and lower values to other layers of
concepts and statements with more-or-less equal weight with respect
to being fundamental to the ontology.
As statements are usually not given a URI explicitly
in RDF files (one has to use ``RDF reification'' to give a triple a URI),
to bind a triple or a set of triples to a level,
one will probably be better of by using named graphs.
Reification is a solution, but a rather verbose and unusual one.

\subsection{Different arrow heads}
To address the problem described in Section~\ref{sec:dia},
we notice that: 1) the edges are often small, so using line styles
(such as dashed, dotted or full) is likely to yield bad results often.
The use of colored edges is an option, but given that many
journals require images to be black and white, it is not suitable
as the standard at least at the moment.
Now different arrow heads are in fact used (almost as a standard) 
as visual cues
to differentiate relations that are yield by properties from those
that associate a class to its superclass,
This is an indicative that using different arrow heads might
be a good strategy to reaching a good solution.
Also, the start of the edge might be a good visual cue to
discern the different type of relations.
We should recall here that in OWL a property might be specified to
satisfy a number of axioms, which care also prone to be represented
with the same artifices.

\subsection{OWL class as subclass of SKOS concept}\label{owlSkos}
There are different
strategies for making an OWL ontology and a SKOS vocabulary compatible~\citep{owlSkos}.
The solutions that are fit for general purposes, among the ones that were found or conceived by the author, are:
\begin{itemize}
	\item describe the ontology as OWL then translate it to SKOS, probably automated by means of a script.
	\item State that each OWL class is also a subclass of the \texttt{skos:Concept}.
		This yields a less redundant result, but it might not be in accordance to SKOS and OWL specifications
		or entail OWL Full.
		For example, the use of some SKOS properties might imply that the subject is an individual of the \texttt{skos:Concept} class.
		This solution needs to be further studied and discussed with specialized groups in order to be used with confidence.
\end{itemize}

The problem with the other solutions:
\begin{itemize}
	\item yields OWL Full (if a SKOS concept, an instance, is also an OWL class); or
	\item yields annotations on instances of OWL classes, not on the classes
		(if we annotate instances of OWL classes with SKOS properties); or
	\item yields a less formal ontology (if we derive the ontology from the vocabulary); or
	\item extrapolates the purpose of SKOS
		(giving enough expressive power to SKOS to achieve an OWL without degeneration of the conceptualization).
\end{itemize}

\section{CONCLUSIONS AND FURTHER WORK}\label{sec:con}
This article is a formalization of the issues to allow
the author and collaborators to confirm (or refute) that these problems
and the solutions herein proposed are sound and have not yet received
solutions by the semantic web community.
These issues and solutions are the result of years of dedication
to the subject, including learning by (online and in-person) classes
and designing dozens of OWL ontologies for academic and State purposes~\citep{pnud5}.

Next steps should include:
\begin{itemize}
	\item the pursue of feedback through circulation of this document (e.g. in online forums)
		to enhance the solutions herein proposed and better understand their possible uses (e.g. in reasoners).
	\item The use of these solutions in ontologies, both new ones an other that are already available.
	\item If these issues prove themselves relevant over time, we will consider contributing to the full realization of solutions.
		For example, we might help achieving and maintaining suitable URIs for the properties described in Section~\ref{sec:props}
		or help in the elaboration of convenient arrow heads for the problem described in Section~\ref{sec:dia}.
	\item The confirmation or refutation of the second possible solution to make OWL and SKOS compatible as described in Section~\ref{owlSkos}.
\end{itemize}

\subsection*{\textit{Acknowledgements}}
The author thanks CNPq for the funding received while researching the topic of this article,
the researchers of IFSC/USP and ICMC/USP for the recurrent collaboration in every situation
where we needed directions for investigation.

% ------------------------------------------------------------------------
\begin{thebibliography}{99}
\fontsize{11}{0}\selectfont
\bibitem[Bechhofer \& Miles, 2008]{owlSkos}
	Bechhofer, S., \& Miles, A. (2008). Using OWL and SKOS. Note of the SWD W3C working group.
From \url{https://www.w3.org/2006/07/SWD/SKOS/skos-and-owl/master.html}

\bibitem[Fabbri \& Oliveira, 2016]{losd}
	Fabbri, R., \& de Oliveira, O. N. (2016). Linked Open Social Database. Github repositories, from \url{https://github.com/ttm/linkedOpenSocialDatabase/raw/master/paper.pdf}

\bibitem[Fabbri, 2017]{fabbri1}
Fabbri, R. (2017). Topological stability and textual differentiation in human interaction networks:
		statistical analysis, visualization and linked data. Doctoral thesis.
		From \url{https://github.com/ttm/thesis/raw/master/thesis-rfabbri.pdf}

\bibitem[Fabbri, 2014]{pnud5}
Fabbri, R. (2014). Social participation ontologies and rules to fuel a social participation linked data cloud.
	Technical report for the United Nations Development Program.
		From \url{https://github.com/ttm/pnud5/raw/master/latex/produto.pdf}

\bibitem[Fabbri et al., 2015]{ops}
Fabbri, R., de Luna, R. B., Martins, R. A. P., Amanqui, F. K. M., \& Moreira, D. D. A. (2015). Social Participation Ontology: community documentation, enhancements and use examples. arXiv preprint arXiv:1501.02662.

\bibitem[Hitzler \& Janowicz, 2013]{semApr}
Hitzler, P., \& Janowicz, K. (2013). Linked Data, Big Data, and the 4th Paradigm. Semantic Web, 4(3), 233-235.

\end{thebibliography}
% ------------------------------------------------------------------------

%For papers written in Portuguese or Spanish.

%\begin{center}
%  TITLE IN ENGLISH
%\end{center}

%\def\abstractname{Abstract}%

%\begin{abstract}
%   Abstract in english
%\end{abstract}

%\keywords{\em{Keywords in english}}

\end{document}

### 2017-08-20 latex/article.tex
started article

@@
 
+\usepackage{grffile}
+
 %%-----------------------------------------------EDIT-----------------------------------------------
-\title{NECESSARY ENHANCEMENTS OF LINKED DATA EXPRESSIVENESS FOR ONTOLOGIES}
+\title{Audiovisual Analytics Vocabulary and Ontology (AAVO): initial core and example expansion}
 
@@
     {\rm \begin{tabular}{l} 
-    \textbf{Renato Fabbri}$^{1}$ - {\textnormal renato.fabbri@gmail.com}\\%
-%    \textbf{Marília M. Pisani}$^{2}$ - {\textnormal marilia.m.pisani@gmail.com}\\
-    {\fontsize{11}{0}\selectfont $^{1}$University of São Paulo, Institute of Mathematical and Computer Sciences - São Carlos, SP, Brazil}\vspace*{-0.05cm} \\
+    \textbf{Renato Fabbri}$$ - {\textnormal renato.fabbri@gmail.com}\\%
+    \textbf{Maria Cristina Ferreira de Oliveira}$$ - {\textnormal cristina@icmc.usp.br}\\
+    {\fontsize{11}{0}\selectfont University of São Paulo, Institute of Mathematical and Computer Sciences - São Carlos, SP, Brazil}\vspace*{-0.05cm} \\
 %    {\fontsize{11}{0}\selectfont $^{2}$Federal University of ABC, Centre for Natural Sciences and Humanities - São Paulo, SP, Brazil}\vspace*{-0.05cm}\\
@@
 \begin{abstract}
-	The semantic web has received many contributions of researchers as ontologies which,
-	in this context, i.e. within RDF linked data, are formalized conceptualization that might use
-	different protocols, such as RDFS, OWL DL and OWL FULL.
-	In this article we describe new expressive techniques which were found necessary after
-	elaborating dozens of OWL ontologies for the scientific academy, the State and the civil society.
-	They consist in: 1) stating possible uses a property might have without incurring into axioms or restrictions;
-	2) assigning a level of priority for an element (class, property, triple); 3) correct depiction in diagrams
-	of relations between classes, between individuals which are imperative, and between individuals which are dispensable;
-	4) association between OWL classes and SKOS concepts.
-	We propose specific rules to accomplish these enhancements and exemplify both its use
-	and the difficulties that arise because these techniques are currently not established as basic to the ontology designer.  
+Visual Analytics might be defined as data mining though interactive visual interfaces.
+The field and has received noticeable consideration by researchers and the industry. 
+The literature, however, is complex because it involves many fields of knowledge
+and is considerably recent.
+In this article we describe an initial organization of this knowledge as an OWL ontology
+and a SKOS vocabulary.
+This effort might be useful in many ways that include conceptual considerations,
+software implementations.
+Within the results and discussions, we expose a core and an example expansion
+of the conceptualization, and incorporate design issues that enhance the
+expressive power of the abstraction. 
 \end{abstract}
 
-\keywords{\em{ Semantic web, Linked data, Ontology, RDFS, OWL}}
+\keywords{\em{OWL, SKOS, Semantic web, Visual analytics, Data visualization}}
 
@@
 \section{INTRODUCTION}
+Visual Analytics' usual definition is data mining (or the science of analytical reasoning facilitated)
+through interactive visual interfaces.
+From this definition, one can grasp that at least three fields are directly related to visual analytics:
+data mining, human-computer interaction, and data visualization.
+Each one of these fields are multidisciplinary and known to be considerably complex
+with multiple theories and vast literature.
+
+This work is proposed as an organization of the knowledge related to visual analytics
+as a SKOS vocabulary and an OWL ontology.
+In other words, this document reports in an initial formalized conceptualization
+in terms of linked data recommended technologies.
+The uses this formalization can have includes:
+\begin{itemize}
+	\item introduction of the Visual Analytics subject for non-specialists.
+	\item Expressing a concise overview.
+	\item Queries.
+	\item Relating objects (e.g. data and techniques) inside the field or between the field and other domains.
+	\item Making inference about the concepts and objects to which they are related.
+\end{itemize}
+
+In the present case, where the formalization is done as linked data,
+the conceptualization allows the queries, inferences, relations, etc.
+to be performed also by machines.
+Therefore, for example, a software might be able to relate a dataset
+or analysis methods to visualization techniques to aid a user of
+for automated reporting.
+
+Section~\ref{sec:methods} uncover the methods.
+Section~\ref{sec:res} is dedicated to stating and discussing the results.
+Final remarks and future work are in Section~\ref{sec:con}.
+
+\section{METHODS}\label{sec:methods}
+The methods described here are standard of the semantic web
+to achieve formalized conceptualizations.
+The next sections addresses the subjects very briefly
+and the interested reader can visit the literature
+of the field and will surely find great items,
+including the W3C recommendations~\citep{w3cld,ldb}.
+
+\subsection{The semantic web}
 The semantic web is constituted by data which are linked in the same way
@@
 field, the terms are equivalent.
+The main topics of this article are data visualization (or visual analytics)
+and semantic web.
+Accordingly, all the sections tackle the subject of the semantic web.
 
+\subsection{RDF}
 The semantic web is built using the Resource Description Framework (RDF).
@@
 Accordingly, one can write:
+
 \begin{Verbatim}[fontsize=\footnotesize]
@@
 	<http://example.org/properties/likes> 
-                <http://example.org/concepts/reading> .
+                <http://example.org/concepts/Reading> .
 \end{Verbatim}
@@
 There are many formats to write/serialize RDF data.
-The example above is written in Turtle and we will use this format throughout this document.
+The example above is written in Turtle format and it will be used throughout this document.
 
-One of the most important notions of the semantic web is that of an ontology.
-An ontology, in this context, is a formalized conceptualizations, comprised by
-concepts and relations between the concepts and between the relations themselves.
-In current semantic web, most simple ontologies are written using the RDFS protocol,
-by which one can specify, among other things:
-\begin{itemize}
-	\item concepts;
-	\item properties, which are concepts that are used as predicates and thus relate concepts;
-	\item special relations between concepts that state that one concept is mode general than the other\footnote{This
-		kind of relation is called hypernymy. Examples: mammal is a hypernym of monkey, drink is a hypernym of beer.};
-	\item the subjects and objects that can occurs in a triple where a specific property is a predicate.
-\end{itemize}
+In real settings, when everything is working as recommended,
+each of these URIs (that are URLs!) might be accessed 
+through HTTP to access more triples referring to the URI accessed.
+From the triples above, one would be able to access triples
+describing each of the properties: \texttt{example:properties/name}, \texttt{example:properties/age}, and\\
+\texttt{example:properties/likes},
+and the concept \texttt{example:concepts/Reading}.
+In fact, the triples above would probably be available in the URI/URL:
+\texttt{example:people/mary}.
+One good working example of this is DBPedia~\citep{dbpedia}.
+The process of accessing a URI to find more triples is called \emph{dereferencing the URI} (or simply dereferencing).
 
-One can also write an ontology by using OWL,
-in which all the expressive capabilities of RDFS are
-available, but one can also, among other things:
-\begin{itemize}
-	\item state ``property axioms'', i.e. specify if a property is e.g. reflexive or transitive;
-	\item state ``class restrictions'', i.e. specify if a class instance e.g. necessarily hold a relation to another class instance or data.
-\end{itemize}
-OWL has a richer vocabulary than RDFS and is (way more) complex.
-This complexity is a drawback together with the greater computational cost
-for performing inference.
-The advantage is the greater power to represent conceptualizations.
+\subsection{RDFS}
+The Resource Description Framework Schema (RDF Schema or simply RDFS)
+is a set of classes and properties for the RDF data model that allows
+basic descriptions of ontologies.
+It supports taxonomic relations (hypernymy, i.e. relations stating that a concept is more general than another),
+bindings of properties to objects and datatypes, and notes (label, comment, see also, is defined by).
 
-The semantic web practices evolves by constant practice and recommendations written
-by many specialists usually after years of research and discussion.
-Therefore, this document is written in a very humble manner,
-without the pretense of yielding new recommendations abruptly,
-but as a way to report issues and candidate workarounds/solutions
-that emerged by years of intense practice in designing ontologies~\citep{pnud5,fabbri1,losd,ops}.
+\subsection{SKOS}
+The Simple Knowledge Organization System (SKOS)
+is a data model for representing controlled vocabularies.
+SKOS is a W3C recommendation to facilitate publication
+and use of vocabularies and is built upon RDF and RDFS.
+It holds itself a vocabulary for concepts, notation, documentation,
+semantic and mapping relations, and collections.
 
-Section~\ref{sec:prob} is dedicated to exposing some of the problems
-in current RDFS and OWL ontologies, with emphasis on the ones
-to which this document holds proposals to resolve or lessen.
-In Section~\ref{sec:prop} such proposals are described and discussed.
-Final remarks are articulated in Section~\ref{sec:con}.
+\subsection{OWL}
+The Web Ontology Language (OWL) is a language for publishing ontologies on the web.
+While RDFS holds basic relations necessary even for very rudimentary organization
+knowledge and data, OWL is complex and allows one to formalize elaborate conceptualizations.
+Using OWL, an ontology might have properties that are required to satisfy a number or axioms,
+and classes that obey restrictions or e.g. are the result of the union of other classes.
 
-\section{THE PROBLEMS}\label{sec:prob}
-The semantic web technology standards are very appreciated by the researchers, developers and users~\citep{semApr}.
-Even so, there are many issues regarding current state of the art.
-For example, if an individual of a class is found also to be a class,
-it implies OWL Full which is a setting presently regarded as not computationally feasible.
-This document is dedicated to proposals that resolve (or alleviate)
-the problems described in the following subsections.
+\subsection{Interviews with specialists and literature consultation}
+The standard approach to design an ontology, according to the literature,
+is to interview specialists of the field to which the ontology is related,
+or to absorb the established literature, or both.
+This work is being developed using both approaches.
+The second author is a data visualization specialist which was
+interviewed by the first author.
+Also, the first author is acquiring a deeper knowledge in the field.
 
-\subsection{No (standard) way to express potential relations}\label{sec:rel}
-There is no recommendation about how to bind to a property a \emph{possible}
-subject or object.
-Using \texttt{rdfs:doman} and \texttt{rdfs:range} has the consequence that
-\textbf{any} subject or object will have be a member of the class specified by
-range or domain.
-There are, however, many situations where this is not the case.
-For example, one might find or create a property \texttt{example:name}.
-Suppose it is intended to be used to express names of animals,
-say because the person created the property while designing an ontology about animals.
-If that is formalized in the property by \texttt{rdfs:domain},
-it implies that whenever the \texttt{example:name} is used, the subject
-will be inferred (e.g. by an automated reasoner) to be an animal.
-Now suppose someone is describing a tool in RDF and wants to write its name.
-If he/she uses the \texttt{example:name} property, it will be erroneously
-expressing that the tool is an animal.
-Such a person is required to create another property \texttt{example2:name}
-to write the names.
-What is worse: if the person wants to create an ontology about tools,
-and wants to express that tools are likely, but not required, to have names,
-the property \texttt{exampleX:name} with domain tool will have the same problem
-as above, but now with tools instead of animals.
-If \textbf{all} the tools have, without exception, a name, then one can express
-it as an OWL existential class restriction.
-But, again, that is often not the case. Example: algorithms are likely to have names,
-but not all of them have names.
-One workaround to this problem is to describe these potential relations as notes to the
-classes and properties, but it does not favor the machine readable design of the semantic web.
+\section{RESULTS AND DISCUSSION}\label{sec:res}
+\subsection{AAVO core}
+\begin{figure}[!htbp] %h or !htbp
+\vspace{-2pt}
+\begin{center}
+% \includegraphics[height=6.7cm,width=9cm]{../figs/aavo0.01_minimum.png}
+\includegraphics[width=\textwidth]{../figs/aavo0.01_minimum.png}
+    \caption{}
+\label{fig:minimum}%
+\end{center}
+\end{figure}
 
-\subsection{All classes and properties are equally fundamental}\label{sec:fun}
-The relevance of this issue can be grasped by
-tacking the expression of conceptualizations for humans.
-Say someone wrote an ontology about machine learning and that it is very complete
-in the sense that it contains many concepts, all those that the author could find.
-It is straightforward to acknowledge that it will be very difficult for a newcomer
-to make sense of everything.
-Also, in many cases it is difficult for anyone to make sense of an ontology when it
-comprises many concepts and relations.
-Thus, it is very useful to have a standard way to express that some set of
-concepts or triples are more fundamental then others in a layered fashion,
-but there isn't a standard way.
+% potential enhancements, from the email to cristina
 
-\subsection{Distinction in diagrams of relations between classes and individuals and between imperative and optional relations}\label{sec:dia}
-This is not a problem found in formalizing the ontology,
-but a problem that arises when depicting the ontology as a diagram.
-For example, the (class of) dinosaurs were succeeded by the (class of) mammals,
-this is a relation between classes.
-A specific animal might have siblings, this is a relation between individuals.
-Also, a relation might be imperative for a class (i.e. it might be expressed as a class restriction)
-or optional.
-These are all poorly distinguished in diagrams that are meant to represent the ontologies.
-
-\subsection{No obvious way to associate an OWL ontology and a SKOS vocabulary}
-In a SKOS vocabulary, a concept is an individual of the class \texttt{skos:concept}.
-In an OWL ontolgy, a concept is a class.
-As stated in the beginning of this section, 
-if an individual of a class is also a class,
-one has an OWL Full context, which is
-currently considered to require inference that is not computable.
-This implies that one has to choose between making
-an OWL ontology or a SKOS vocabulary or find a way to
-make them compatible.
-This might be the most complex issue tackled in this document.
-
-\section{ENHANCEMENT PROPOSALS}\label{sec:prop}
-As the problems stated in the previous section are distinct,
-the proposed enhancements are described case-by-case.
-
-\subsection{Potential domain, range, predicate and post predicate}
-To address the problem described in Section~\ref{sec:rel},
-we propose creating a new set of properties:
-\begin{itemize}
-	\item potential/nonrestrictive domain and range of a property might be expressed using\\
-		\texttt{example:potentialDomain} and \texttt{example:potentialRange}.
-	\item potential predicates when an instance of a class is a subject or an object might
-		be expressed using\\
-		\texttt{example:potentialPredicate} and \texttt{example:potentialPostPredicate}.
-\end{itemize}
-
-The \texttt{example:} in these triples should be replaced by a true namespace prefix, such
-as \texttt{http://purl.org/socialparticipation/rdfe/}.
-This was not done until the moment because we chose to wait a little longer,
-present this article and distribute it online to find out if there is already some
-solution currently in use and if there is some trustworthy maintainer that is willing
-to keep these properties online in a more suitable namespace.
-
-\subsection{Ascribing numeric values that reflect how fundamental is something}\label{sec:props}
-To address the problem described in Section~\ref{sec:fun},
-every property and class might me a subject to a triple:
-\begin{Verbatim}[fontsize=\footnotesize]
-	example:something example:fundamental 4 .
-\end{Verbatim}
-\noindent with ``4'' replaced by any other numerical value.
-At first, it seems reasonable to assume that these values are
-valid within the ontology and relative to each other.
-Accordingly, one might ascribe \texttt{fundamental 100}
-some kernel of meaning and lower values to other layers of
-concepts and statements with more-or-less equal weight with respect
-to being fundamental to the ontology.
-As statements are usually not given a URI explicitly
-in RDF files (one has to use ``RDF reification'' to give a triple a URI),
-to bind a triple or a set of triples to a level,
-one will probably be better of by using named graphs.
-Reification is a solution, but a rather verbose and unusual one.
-
-\subsection{Different arrow heads}
-To address the problem described in Section~\ref{sec:dia},
-we notice that: 1) the edges are often small, so using line styles
-(such as dashed, dotted or full) is likely to yield bad results often.
-The use of colored edges is an option, but given that many
-journals require images to be black and white, it is not suitable
-as the standard at least at the moment.
-Now different arrow heads are in fact used (almost as a standard) 
-as visual cues
-to differentiate relations that are yield by properties from those
-that associate a class to its superclass,
-This is an indicative that using different arrow heads might
-be a good strategy to reaching a good solution.
-Also, the start of the edge might be a good visual cue to
-discern the different type of relations.
-We should recall here that in OWL a property might be specified to
-satisfy a number of axioms, which care also prone to be represented
-with the same artifices.
-
-\subsection{OWL class as subclass of SKOS concept}\label{owlSkos}
-There are different
-strategies for making an OWL ontology and a SKOS vocabulary compatible~\citep{owlSkos}.
-The solutions that are fit for general purposes, among the ones that were found or conceived by the author, are:
-\begin{itemize}
-	\item describe the ontology as OWL then translate it to SKOS, probably automated by means of a script.
-	\item State that each OWL class is also a subclass of the \texttt{skos:Concept}.
-		This yields a less redundant result, but it might not be in accordance to SKOS and OWL specifications
-		or entail OWL Full.
-		For example, the use of some SKOS properties might imply that the subject is an individual of the \texttt{skos:Concept} class.
-		This solution needs to be further studied and discussed with specialized groups in order to be used with confidence.
-\end{itemize}
-
-The problem with the other solutions:
-\begin{itemize}
-	\item yields OWL Full (if a SKOS concept, an instance, is also an OWL class); or
-	\item yields annotations on instances of OWL classes, not on the classes
-		(if we annotate instances of OWL classes with SKOS properties); or
-	\item yields a less formal ontology (if we derive the ontology from the vocabulary); or
-	\item extrapolates the purpose of SKOS
-		(giving enough expressive power to SKOS to achieve an OWL without degeneration of the conceptualization).
-\end{itemize}
+\subsection{AAVO example expansion}
+\begin{figure}[!htbp] %h or !htbp
+\vspace{-2pt}
+\begin{center}
+\includegraphics[width=\textwidth]{../figs/aavo0.01_exemplified.png}
+    \caption{}
+\label{fig:minimum}%
+\end{center}
+\end{figure}
 
 \section{CONCLUSIONS AND FURTHER WORK}\label{sec:con}
-This article is a formalization of the issues to allow
-the author and collaborators to confirm (or refute) that these problems
-and the solutions herein proposed are sound and have not yet received
-solutions by the semantic web community.
-These issues and solutions are the result of years of dedication
-to the subject, including learning by (online and in-person) classes
-and designing dozens of OWL ontologies for academic and State purposes~\citep{pnud5}.
-
-Next steps should include:
-\begin{itemize}
-	\item the pursue of feedback through circulation of this document (e.g. in online forums)
-		to enhance the solutions herein proposed and better understand their possible uses (e.g. in reasoners).
-	\item The use of these solutions in ontologies, both new ones an other that are already available.
-	\item If these issues prove themselves relevant over time, we will consider contributing to the full realization of solutions.
-		For example, we might help achieving and maintaining suitable URIs for the properties described in Section~\ref{sec:props}
-		or help in the elaboration of convenient arrow heads for the problem described in Section~\ref{sec:dia}.
-	\item The confirmation or refutation of the second possible solution to make OWL and SKOS compatible as described in Section~\ref{owlSkos}.
-\end{itemize}
 
 \subsection*{\textit{Acknowledgements}}
-The author thanks CNPq for the funding received while researching the topic of this article,
+The authors thank CNPq and FAPESP for the fundings received while researching the topic of this article,
 the researchers of IFSC/USP and ICMC/USP for the recurrent collaboration in every situation

### 2017-08-21 latex/article.tex
finished first draft

@@
 
-\section{INTRODUCTION}
+\section{INTRODUCTION}\label{sec:intro}
 Visual Analytics' usual definition is data mining (or the science of analytical reasoning facilitated)
@@
 to achieve formalized conceptualizations.
-The next sections addresses the subjects very briefly
+The next sections address the subjects very briefly
 and the interested reader can visit the literature
@@
 \texttt{example:people/mary}.
-One good working example of this is DBPedia~\citep{dbpedia}.
+One good working example\footnote{Visit \url{http://dbpedia.org/page/Rhesus_macaque}
+and click on the concepts and properties to start browsing the web of linked data.}
+of this is DBPedia~\citep{dbpedia}.
 The process of accessing a URI to find more triples is called \emph{dereferencing the URI} (or simply dereferencing).
@@
 \section{RESULTS AND DISCUSSION}\label{sec:res}
-\subsection{AAVO core}
+% a word about AAVO, audiovisual, discerning core and expansions
+% main concepts
+Using the framework exposed in the previous section,
+we elaborated an initial vocabulary and ontology for
+visual analytics: AAVO (Audiovisual Analytics Vocabulary and Ontology).
+The inclusion of ``audio'' is both a reminder of the possibilities
+available for using audio to represent data and perceive patterns,
+and a fitting incorporation of audio to visual analytics given
+audiovisual capabilites of current ordinary computers.
+
+The main concepts and their interrelation are presented in
+Section~\ref{sec:core} while an example extension is on Section~\ref{sec:ext}.
+Section~\ref{sec:voc} hold annotations for the vocabulary which are not promptly
+given by the previous sections.
+Some of the relations bellow were expressed by very recent 
+techniques that are described in~\cite{enhance}.
+Their meaning, though, might be easily inferred.
+
+\subsection{AAVO core}\label{sec:core}
+The core of AAVO is designed to be minimal and hold the following concepts as
+depicted in Figure~\ref{fig:minimum}:
+\begin{itemize}
+	\item Visualization: a technique to generate a Visual Representation from Data.
+	\item Visual Representation: a representation of data by visual cues.
+		A Visual Representation can be an Image or an Animation.
+	\item Data: a set o values be them qualitative or quantitative~\citep{wikipData}.
+	\item Dataset Type: a type of organization and meaning of data~\cite{munzner}.
+	\item Processing: transforms Data into Data. Pre-processing is a kind of Processing.
+\end{itemize}
+
+We envision that there should be at least the following concepts in
+AAVO core when it reaches maturity:
+\begin{itemize}
+	\item Hypothesis: a proposed explanation for a phenomenon that might be
+		1) given beforehand for being proved or refuted by an Analysis,
+		2) shaed by means of an Analysis,
+		or 3) presented by means of a Visualization.
+	\item Analysis: a set of procedures used to obtain understanding about
+		Data or a phenomenon.
+	\item Task/Purpose/Application: the goal or objective of an Analysis.
+\end{itemize}
+\noindent These concepts were not included in AAVO core (e.g. Figure~\ref{fig:minimum})
+because we chose to better conceptualize the field in order to avoid too much rework.
+From the definitions above, raises the question:
+should we also include Phenomenon among these core concepts?
+
+Other relations that can be added to the core or to an extension but
+are directly related to the core:
+\begin{itemize}
+	\item Visualization is a type of Processing.
+	\item Visual Representation is a Dataset Type.
+	\item Processing ``suitable for'' Data.
+	\item Visualization ``number of dimensions'' real (not double as stated for now).
+\end{itemize}
+
+An example of question still left unanswered:
+is a Visualization only outputs Visual Representation
+or can it output other Data?
+This and many other questions might have a resolution
+that are genuinely dependent on the conceptual design of the ontology.
+
 \begin{figure}[!htbp] %h or !htbp
@@
 \includegraphics[width=\textwidth]{../figs/aavo0.01_minimum.png}
-    \caption{}
+	\caption{AAVO core, discussed in Section~\ref{sec:core}.
+	A Visualization outputs a Visual Representation that can be an Image or an Animation.
+	A Visualization is suitable for a Dataset Type and for a Processing routine which transforms
+	Data into Data.}
 \label{fig:minimum}%
@@
 
-\subsection{AAVO example expansion}
+\subsection{AAVO example expansion}\label{sec:ext}
+There are many ways in which the AAVO core might be expanded.
+Figure~\ref{fig:exemp} is an example expansion.
+Concepts were added which are hyponyms to Dataset Type (Temporal Series, Relational Data),
+to Pre-Processing (Z-Score, Cleaning), Processing (MDS, Statistical Test)
+and Visualization (Heat Map, Histogram, Scatter Plot, Timeline).
+Some example of further subclasses are also added.
+A different kind of expansion was achieved by including (Data) Availability
+and the less general concepts of Dynamic Availability and Static Availability.
+A Graph is regarded as a Network without any context or further attributes beyond
+nodes and edges.
+
+Ideally, AAVO expansions should reach related fields, such as HCI,
+by linking to other existing ontologies (such as DBPedia) or incorporating
+enough concepts to then bind and rely in third party conceptualizations.
+
 \begin{figure}[!htbp] %h or !htbp
@@
 \includegraphics[width=\textwidth]{../figs/aavo0.01_exemplified.png}
-    \caption{}
-\label{fig:minimum}%
+    \caption{Added less general concepts: Statistical Test, MDS, Timeline, Z-Score, Network, etc.
+	A thorough consideration of this expansion of the AAVO core is at Section~\ref{sec:ext}.}
+\label{fig:exemp}%
 \end{center}
@@
 
+\subsection{Vocabulary annotations}\label{sec:voc}
+Beyond what is made explicit in the previous sections,
+there are some aspects of the knowledge and language that
+are to be directly added to the SKOS vocabulary.
+Examples:
+\begin{itemize}
+	\item in a dataset, an element is also called: an item, an observation, an individual, a point,
+		and even a data point and a data row. 
+	\item A graph node is also called: a vertex, and every name that are used to designate an element.
+	\item A graph edge is also called: a link, a bond, a line, and a connection.
+	\item Z-scores are also called: standard scores, normal scores, standardized variables, and z-values. 
+\end{itemize}
+
 \section{CONCLUSIONS AND FURTHER WORK}\label{sec:con}
+This initial formalized conceptualization of the AAVO
+holds some relations which are not explicitly described by
+current literature mainly because:
+1) AAVO was created with the purpose of reaching a sound
+conceptualization that allows it to be formalized as linked data;
+2) AAVO has the purpose of representing the knowledge in Visual Analytics
+to enable inference by machines.
+There are other uses for AAVO which were uncovered in Section~\ref{sec:intro},
+but for those there are conceptual models available~\cite{munzner}.
+
+Potential further steps include:
+\begin{itemize}
+	\item the inclusion of the concepts Hypothesis, Analysis, and Task
+		into the AAVO Core.
+	\item Realizing AAVO expansions until the reached concepts can be linked
+		to other ontologies that are sound, used and maintained.
+	\item Using AAVO for obtaining interesting relations by means of automated
+		inference and for assisting a (audio) visual analytics software.
+\end{itemize}
 
 \subsection*{\textit{Acknowledgements}}
-The authors thank CNPq and FAPESP for the fundings received while researching the topic of this article,
+The authors thank CNPq and FAPESP for the funding received while researching the topic of this article,
 the researchers of IFSC/USP and ICMC/USP for the recurrent collaboration in every situation
@@
 \fontsize{11}{0}\selectfont
-\bibitem[Bechhofer \& Miles, 2008]{owlSkos}
-	Bechhofer, S., \& Miles, A. (2008). Using OWL and SKOS. Note of the SWD W3C working group.
-From \url{https://www.w3.org/2006/07/SWD/SKOS/skos-and-owl/master.html}
+\bibitem[W3C, 2010]{w3cld}
+	W3C (2017). LINKED DATA CURRENT STATUS, from \url{https://www.w3.org/standards/techs/linkeddata}
+
+\bibitem[Heath \& Bizer, 2011]{ldb}
+	Heath, T. \& Bizer, C. (2011). Linked Data: Evolving the Web into a Global Data Space (1st edition). Synthesis Lectures on the Semantic Web: Theory and Technology, 1:1, 1-136. Morgan \& Claypool.
+
+\bibitem[Lehmann et al., 2015]{dbpedia}
+	Lehmann, J., Isele, R., Jakob, M., Jentzsch, A., Kontokostas, D., Mendes, P.
+N., ...
+\& Bizer, C.
+(2015).
+DBpedia–a large-scale, multilingual knowledge base extracted from Wikipedia.
+Semantic Web, 6(2), 167-195.
+
+\bibitem[Fabbri, 2017]{enhance}
+	Fabbri, R. (2017). Necessary enhancements of linked data expressiveness for ontologies.
+		Encontro Nacional de Modelagem Computacional 2017 (XX ENMC).
+		From \url{https://github.com/ttm/ontologyEnhancements/raw/master/article.pdf}
+
+\bibitem[Wikipedia, 2017]{wikipData}
+	Data. (2017, August 21). In Wikipedia, The Free Encyclopedia. Retrieved
+		22:31, August 21, 2017
+		, from \url{https://en.wikipedia.org/w/index.php?title=Data&oldid=796493851}
+
+\bibitem[Munzner, 2014]{munzner}
+	Munzner, T. (2014). Visualization analysis and design. CRC press.
 

### 2017-08-21 src/ontology.py

@@ -41,7 +41,8 @@ triples = [
         (aao.Histogram, owl.subClassOf, aao.Visualization),
         (aao.Histogram, aao.suitableFor, aao.StatisticalTest),
         (aao.ScatterPlot, owl.subClassOf, aao.Visualization),
-        (aao.Timeline, owl.subClassOf, aao.ScatterPlot), # Ok?? Enhance this
+        (aao.Timeline, owl.subClassOf, aao.Visualization), # Ok?? Enhance this
+        (aao.Timeline, owl.suitableFor, aao.TemporalSeries), # Ok?? Enhance this
         (aao.ScatterPlot, aao.suitableFor, aao.MDS),
 
         (aao.Visualization, aao.output, aao.VisualRepresentation), # put Visual Representation as a subclass of Data?

### 2017-08-21 src/ontologyMin.py

@@ -6,17 +6,17 @@ aao = p.rdf.NS.aao # Audiovisual Analytics Ontology
 a =  p.rdf.NS.rdf.type
 
 triples = [
-        (aao.Data, aao.type, aao.Datatype),
+        (aao.Data, aao.type, aao.DatasetType),
 
         (aao.Processing, aao.input, aao.Data),
         (aao.Processing, aao.output, aao.Data),
-        (aao.Processing, aao.suitableFor, aao.Datatype),
+        (aao.Processing, aao.suitableFor, aao.DatasetType),
 
         (aao.PreProcessing, owl.subClassOf, aao.Processing),
 
         (aao.Visualization, aao.uses, aao.Processing), # better term than uses?
         (aao.Visualization, aao.suitableFor, aao.Processing),
-        (aao.Visualization, aao.suitableFor, aao.Datatype),
+        (aao.Visualization, aao.suitableFor, aao.DatasetType),
         (aao.Visualization, aao.input, aao.Data),
         (aao.Visualization, aao.numberOfDimensions, xsd.double),
 
@@ -44,5 +44,5 @@ o.render(path, "aavo0.01_minimum")
 # Notes:
 # * Include Audio(visualization)
 # * Make an even simpler core
-# * Main concepts: Visualization, Analysis*, Data, Datatype, Processing, PreProcessing, Hypothesis, Task/Purpose/Application
+# * Main concepts: Visualization, Analysis*, Data, DatasetType, Processing, PreProcessing, Hypothesis, Task/Purpose/Application
 # * Consider existential and universal restrictions

### 2017-08-21 latex/article.tex
ordered the bib items

@@
 \fontsize{11}{0}\selectfont
-\bibitem[W3C, 2010]{w3cld}
-	W3C (2017). LINKED DATA CURRENT STATUS, from \url{https://www.w3.org/standards/techs/linkeddata}
+\bibitem[Fabbri, 2017]{enhance}
+	Fabbri, R. (2017). Necessary enhancements of linked data expressiveness for ontologies.
+		Encontro Nacional de Modelagem Computacional 2017 (XX ENMC).
+		From \url{https://github.com/ttm/ontologyEnhancements/raw/master/article.pdf}
 
@@
 
-\bibitem[Fabbri, 2017]{enhance}
-	Fabbri, R. (2017). Necessary enhancements of linked data expressiveness for ontologies.
-		Encontro Nacional de Modelagem Computacional 2017 (XX ENMC).
-		From \url{https://github.com/ttm/ontologyEnhancements/raw/master/article.pdf}
+\bibitem[Munzner, 2014]{munzner}
+	Munzner, T. (2014). Visualization analysis and design. CRC press.
+
+\bibitem[W3C, 2010]{w3cld}
+	W3C (2017). LINKED DATA CURRENT STATUS, from \url{https://www.w3.org/standards/techs/linkeddata}
 
@@
 		, from \url{https://en.wikipedia.org/w/index.php?title=Data&oldid=796493851}
-
-\bibitem[Munzner, 2014]{munzner}
-	Munzner, T. (2014). Visualization analysis and design. CRC press.
-
-\bibitem[Fabbri \& Oliveira, 2016]{losd}
-	Fabbri, R., \& de Oliveira, O. N. (2016). Linked Open Social Database. Github repositories, from \url{https://github.com/ttm/linkedOpenSocialDatabase/raw/master/paper.pdf}
-
-\bibitem[Fabbri, 2017]{fabbri1}
-Fabbri, R. (2017). Topological stability and textual differentiation in human interaction networks:
-		statistical analysis, visualization and linked data. Doctoral thesis.
-		From \url{https://github.com/ttm/thesis/raw/master/thesis-rfabbri.pdf}
-
-\bibitem[Fabbri, 2014]{pnud5}
-Fabbri, R. (2014). Social participation ontologies and rules to fuel a social participation linked data cloud.
-	Technical report for the United Nations Development Program.
-		From \url{https://github.com/ttm/pnud5/raw/master/latex/produto.pdf}
-
-\bibitem[Fabbri et al., 2015]{ops}
-Fabbri, R., de Luna, R.
-B., Martins, R.
-A.
-P., Amanqui, F.
-K.
-M., \& Moreira, D.
-D.
-A.
-(2015).
-Social Participation Ontology: community documentation, enhancements and use examples.
-arXiv preprint arXiv:1501.02662.
-
-\bibitem[Hitzler \& Janowicz, 2013]{semApr}
-Hitzler, P., \& Janowicz, K. (2013). Linked Data, Big Data, and the 4th Paradigm. Semantic Web, 4(3), 233-235.
 

### 2017-08-22 latex/article.tex
finished review

@@
 \usepackage{grffile}
+\usepackage{afterpage}
 
@@
 Visual Analytics might be defined as data mining though interactive visual interfaces.
-The field and has received noticeable consideration by researchers and the industry. 
+The field and has received prominent consideration by researchers, developers and the industry. 
 The literature, however, is complex because it involves many fields of knowledge
@@
 and a SKOS vocabulary.
-This effort might be useful in many ways that include conceptual considerations,
+This effort might be useful in many ways that include conceptual considerations and
 software implementations.
@@
 
-This work is proposed as an organization of the knowledge related to visual analytics
-as a SKOS vocabulary and an OWL ontology.
-In other words, this document reports in an initial formalized conceptualization
+This work reports an initial organization of the knowledge related to visual analytics
+as a SKOS vocabulary and an OWL ontology, i.e. a formalized conceptualization
 in terms of linked data recommended technologies.
-The uses this formalization can have includes:
+The uses this formalization can have include:
 \begin{itemize}
@@
 
-In the present case, where the formalization is done as linked data,
+In the present case, where the formalization is realized as linked data,
 the conceptualization allows the queries, inferences, relations, etc.
 to be performed also by machines.
-Therefore, for example, a software might be able to relate a dataset
-or analysis methods to visualization techniques to aid a user of
+Therefore, for example, a software might relate a dataset
+or analysis methods to visualization techniques in order to assist a user or
 for automated reporting.
 
-Section~\ref{sec:methods} uncover the methods.
-Section~\ref{sec:res} is dedicated to stating and discussing the results.
+Section~\ref{sec:methods} uncovers the methods and technologies
+needed for stating and discussing the results in Section~\ref{sec:res}.
 Final remarks and future work are in Section~\ref{sec:con}.
@@
 to achieve formalized conceptualizations.
-The next sections address the subjects very briefly
-and the interested reader can visit the literature
-of the field and will surely find great items,
-including the W3C recommendations~\citep{w3cld,ldb}.
+Thus, the next sections address the subjects very briefly.
+The interested reader should visit the vanilla literature,
+especially the W3C recommendations~\citep{w3cld,ldb}.
 
 \subsection{The semantic web}
-The semantic web is constituted by data which are linked in the same way
-as web pages are: through HTTP and URLs.
+The semantic web is constituted by data which is linked in the same way
+web pages are: through HTTP and URLs.
 W3C recommendations are the main sources of protocols and best practices of the field.
-In practice, the terms 'linked data' and 'semantic web' are most often used interchangeably.
-A distinction might arise in some contexts where on need to refer to the linked data or
+The terms 'linked data' and 'semantic web' are most often used interchangeably.
+A distinction might arise in some contexts where one needs to refer to the linked data or
 the semantic web created by all or some portion of linked data, but, as a knowledge and technological
@@
 The main topics of this article are data visualization (or visual analytics)
-and semantic web.
-Accordingly, all the sections tackle the subject of the semantic web.
+and semantic web, and all the sections tackle the subject of the semantic web
+for it is the framework in which data visualization knowledge is represented.
 
@@
 The RDF data model is based on making statements in the form of triples
-(``subject-predicate-object'') and using Unique Resource Identifier for
+(``subject-predicate-object'') and using Unique Resource Identifiers (URIs) for
 objects and concepts.
@@
 \end{Verbatim}
-\noindent to express that there is person called Mary Shastacian which is 57 years old and likes reading.
+\noindent to express that there is a 57 years old person called Mary Shastacian which likes reading.
 There are many formats to write/serialize RDF data.
-The example above is written in Turtle format and it will be used throughout this document.
+The example above is written in Turtle and it will be the format used throughout this document.
 
@@
 each of these URIs (that are URLs!) might be accessed 
-through HTTP to access more triples referring to the URI accessed.
+through HTTP to reach more triples referring to the URI.
 From the triples above, one would be able to access triples
@@
 and use of vocabularies and is built upon RDF and RDFS.
-It holds itself a vocabulary for concepts, notation, documentation,
+It is itself a vocabulary for concepts, notation, documentation,
 semantic and mapping relations, and collections.
@@
 we elaborated an initial vocabulary and ontology for
-visual analytics: AAVO (Audiovisual Analytics Vocabulary and Ontology).
+visual analytics: the AAVO (Audiovisual Analytics Vocabulary and Ontology).
 The inclusion of ``audio'' is both a reminder of the possibilities
 available for using audio to represent data and perceive patterns,
-and a fitting incorporation of audio to visual analytics given
+and a desirable incorporation of audio to visual analytics given
 audiovisual capabilites of current ordinary computers.
 
-The main concepts and their interrelation are presented in
+The main concepts and their interrelations are presented in
 Section~\ref{sec:core} while an example extension is on Section~\ref{sec:ext}.
-Section~\ref{sec:voc} hold annotations for the vocabulary which are not promptly
+Section~\ref{sec:voc} holds annotations for the vocabulary which are not promptly
 given by the previous sections.
-Some of the relations bellow were expressed by very recent 
+Some of the relations bellow are expressed using very recent 
 techniques that are described in~\cite{enhance}.
@@
 	\item Visualization: a technique to generate a Visual Representation from Data.
-	\item Visual Representation: a representation of data by visual cues.
+	\item Visual Representation: a representation of Data by visual cues.
 		A Visual Representation can be an Image or an Animation.
-	\item Data: a set o values be them qualitative or quantitative~\citep{wikipData}.
-	\item Dataset Type: a type of organization and meaning of data~\cite{munzner}.
+	\item Data: a set of values, be them qualitative or quantitative~\citep{wikipData}.
+	\item Dataset Type: a type of organization and meaning of data~\citep{munzner}.
 	\item Processing: transforms Data into Data. Pre-processing is a kind of Processing.
 \end{itemize}
-
-We envision that there should be at least the following concepts in
-AAVO core when it reaches maturity:
-\begin{itemize}
-	\item Hypothesis: a proposed explanation for a phenomenon that might be
-		1) given beforehand for being proved or refuted by an Analysis,
-		2) shaed by means of an Analysis,
-		or 3) presented by means of a Visualization.
-	\item Analysis: a set of procedures used to obtain understanding about
-		Data or a phenomenon.
-	\item Task/Purpose/Application: the goal or objective of an Analysis.
-\end{itemize}
-\noindent These concepts were not included in AAVO core (e.g. Figure~\ref{fig:minimum})
-because we chose to better conceptualize the field in order to avoid too much rework.
-From the definitions above, raises the question:
-should we also include Phenomenon among these core concepts?
-
-Other relations that can be added to the core or to an extension but
-are directly related to the core:
-\begin{itemize}
-	\item Visualization is a type of Processing.
-	\item Visual Representation is a Dataset Type.
-	\item Processing ``suitable for'' Data.
-	\item Visualization ``number of dimensions'' real (not double as stated for now).
-\end{itemize}
-
-An example of question still left unanswered:
-is a Visualization only outputs Visual Representation
-or can it output other Data?
-This and many other questions might have a resolution
-that are genuinely dependent on the conceptual design of the ontology.
 
@@
 
+We envision that there should be at least the following concepts in
+AAVO core when it reaches maturity:
+\begin{itemize}
+	\item Hypothesis: a proposed explanation for a phenomenon that might be
+		1) given beforehand and amenable to being proved or refuted by an Analysis,
+		2) shared by means of an Analysis,
+		or 3) presented by means of a Visualization.
+	\item Analysis: a set of procedures used to obtain understanding about
+		Data or a phenomenon.
+	\item Task/Purpose/Application: the goal or objective of an Analysis.
+\end{itemize}
+\noindent These concepts were not included in AAVO core (e.g. Figure~\ref{fig:minimum})
+because we still deciding on the best way to do so.
+From the definitions above, raises the question:
+should we also include Phenomenon among these core concepts?
+
+Other relations that can be added to the core (or to an extension, but
+are directly related to the core):
+\begin{itemize}
+	\item Visualization is a type of Processing.
+	\item Visual Representation is a Dataset Type.
+	\item Processing ``suitable for'' Data.
+	\item Visualization ``number of dimensions'' real (not double as stated for now).
+\end{itemize}
+
+An example of question still left unanswered:
+a Visualization only outputs Visual Representation
+or can it output other Data(set Type)?
+This and many other questions might have a resolution
+that are genuinely dependent on the conceptual design of the ontology.
+
 % potential enhancements, from the email to cristina
@@
 and Visualization (Heat Map, Histogram, Scatter Plot, Timeline).
-Some example of further subclasses are also added.
+Some examples of further subclasses are also added.
 A different kind of expansion was achieved by including (Data) Availability
 and the less general concepts of Dynamic Availability and Static Availability.
-A Graph is regarded as a Network without any context or further attributes beyond
+A Graph is regarded as a bare Network without any context or further attributes beyond
 nodes and edges.
-
-Ideally, AAVO expansions should reach related fields, such as HCI,
-by linking to other existing ontologies (such as DBPedia) or incorporating
-enough concepts to then bind and rely in third party conceptualizations.
 
@@
 \includegraphics[width=\textwidth]{../figs/aavo0.01_exemplified.png}
-    \caption{Added less general concepts: Statistical Test, MDS, Timeline, Z-Score, Network, etc.
-	A thorough consideration of this expansion of the AAVO core is at Section~\ref{sec:ext}.}
+    \caption{Example incorporation of less general concepts to AAVO: Statistical Test, MDS, Timeline, Z-Score, Network, etc.
+	A thorough consideration of this expansion of the AAVO core is at Section~\ref{sec:ext}.
+	In case this image is not being properly visualized in paper,
+	an online PDF should be at: \url{https://github.com/ttm/aavo/raw/master/latex/article.pdf}.}
 \label{fig:exemp}%
@@
 \end{figure}
+
+Ideally, AAVO expansions should reach related fields, such as HCI,
+by linking to other existing ontologies (such as DBPedia) or incorporating
+enough concepts to then bind and rely in third party conceptualizations.
 
@@
 holds some relations which are not explicitly described by
-current literature mainly because:
-1) AAVO was created with the purpose of reaching a sound
-conceptualization that allows it to be formalized as linked data;
-2) AAVO has the purpose of representing the knowledge in Visual Analytics
+current literature mainly because of the purposes:
+1) of reaching a sound
+conceptualization that allows a formalization as linked data;
+2) of representing the knowledge in Visual Analytics
 to enable inference by machines.
-There are other uses for AAVO which were uncovered in Section~\ref{sec:intro},
-but for those there are conceptual models available~\cite{munzner}.
+There are other uses for AAVO, uncovered in Section~\ref{sec:intro},
+for which there are conceptual models available~\citep{munzner,ward}.
 
@@
 	\item the inclusion of the concepts Hypothesis, Analysis, and Task
-		into the AAVO Core.
+		into the AAVO core.
 	\item Realizing AAVO expansions until the reached concepts can be linked
@@
 	\item Using AAVO for obtaining interesting relations by means of automated
-		inference and for assisting a (audio) visual analytics software.
+		inference and for assisting a (audio)visual analytics software.
 \end{itemize}
@@
 \bibitem[Fabbri, 2017]{enhance}
-	Fabbri, R. (2017). Necessary enhancements of linked data expressiveness for ontologies.
+	Fabbri, R. (2017). Enhancements of linked data expressiveness for ontologies.
 		Encontro Nacional de Modelagem Computacional 2017 (XX ENMC).
@@
 	W3C (2017). LINKED DATA CURRENT STATUS, from \url{https://www.w3.org/standards/techs/linkeddata}
+
+\bibitem[Ward et al., 2010]{ward}
+	Ward, M. O., Grinstein, G., \& Keim, D. (2010). Interactive data visualization: foundations, techniques, and applications. CRC Press.
 

### 2017-08-29 latex/article.tex
incorporadas contribs da Cristina

@@
 \begin{abstract}
-Visual Analytics might be defined as data mining though interactive visual interfaces.
-The field and has received prominent consideration by researchers, developers and the industry. 
-The literature, however, is complex because it involves many fields of knowledge
+Visual Analytics might be defined as data mining assisted by interactive visual interfaces.
+The field has been receiving prominent consideration by researchers, developers and the industry. 
+The literature, however, is complex because it involves multiple fields of knowledge
 and is considerably recent.
-In this article we describe an initial organization of this knowledge as an OWL ontology
+In this article we describe an initial tentative organization of the knowledge in the field as an OWL ontology
 and a SKOS vocabulary.
@@
 \section{INTRODUCTION}\label{sec:intro}
-Visual Analytics' usual definition is data mining (or the science of analytical reasoning facilitated)
-through interactive visual interfaces.
+Visual Analytics is often defined as data mining (or the science of analytical reasoning)  facilitated by interactive visual interfaces~\citep{wikiVA,visMaster}.
+% * <cristina@icmc.usp.br> 2017-08-24T21:25:25.777Z:
+% 
+% > Visual Analytics is often defined as data mining (or the science of analytical reasoning)  facilitated by interactive visual interfaces
+% 
+% não seria o caso de inserir refs que suportem a afirmação?
+% 
+% ^ <renato.fabbri@gmail.com> 2017-08-29T15:03:18.752Z:
+% 
+% feito
+%
+% ^.
 From this definition, one can grasp that at least three fields are directly related to visual analytics:
 data mining, human-computer interaction, and data visualization.
-Each one of these fields are multidisciplinary and known to be considerably complex
+Each one of these fields are inherently multidisciplinary and known to be considerably complex,
 with multiple theories and vast literature.
+Researchers entering into the field of visual analytics can easily find themselves confused with such a diverse material, possibly presented using  terminology that is not necessarily consistent across the different fields.
 
-This work reports an initial organization of the knowledge related to visual analytics
+This work reports an initial attempt of organizing the knowledge related to visual analytics
 as a SKOS vocabulary and an OWL ontology, i.e. a formalized conceptualization
+% * <cristina@icmc.usp.br> 2017-08-24T21:34:10.696Z:
+% 
+% > SKOS
+% seria necessário definir os termos? SKOS, OWL... (ou é razoável assumir que a audiência está familiarizada com eles?
+% 
+% ^ <renato.fabbri@gmail.com> 2017-08-25T19:55:32.810Z:
+% 
+% eles não são tão triviais de serem formalizados e, assim, achei por bem definir na seção de métodos...
+%
+% ^.
 in terms of linked data recommended technologies.
-The uses this formalization can have include:
+Such a formalization can have multiple uses, including:
 \begin{itemize}
-	\item introduction of the Visual Analytics subject for non-specialists.
-	\item Expressing a concise overview.
-	\item Queries.
-	\item Relating objects (e.g. data and techniques) inside the field or between the field and other domains.
-	\item Making inference about the concepts and objects to which they are related.
+	\item Facilitating the introduction of the Visual Analytics subject to non-specialists.
+	\item Expressing a concise overview of the field.
+	\item Expressing a domain knowledge against which queries might be performed.
+% * <cristina@icmc.usp.br> 2017-08-24T21:36:14.552Z:
+% 
+% >  Queries
+% esse uso não está expresso usando um verbo, como os demais... do you mean Formulate queries? 
+% 
+% ^ <renato.fabbri@gmail.com> 2017-08-25T19:57:36.800Z:
+% 
+% melhorou?
+%
+% ^.
+	\item Relating objects (e.g. data and techniques) within the field or between the field and other domains.
+	\item Making inferences about the concepts and objects to which they are related.
 \end{itemize}
@@
 to be performed also by machines.
-Therefore, for example, a software might relate a dataset
-or analysis methods to visualization techniques in order to assist a user or
+Therefore, a software might, for example,  relate a dataset
+or analysis methods to specific visualization techniques, in order to assist a user designing a visualization, or
 for automated reporting.
@@
 web pages are: through HTTP and URLs.
-W3C recommendations are the main sources of protocols and best practices of the field.
-The terms 'linked data' and 'semantic web' are most often used interchangeably.
-A distinction might arise in some contexts where one needs to refer to the linked data or
-the semantic web created by all or some portion of linked data, but, as a knowledge and technological
-field, the terms are equivalent.
+W3C recommendations provide the main source of protocols and best practices for the field.
+The terms `linked data' and `semantic web' are most often used interchangeably.
+A distinction might arise in some contexts where one needs to refer to the (linked) data or
+the (semantic) web created by all or some portion of linked data, but, generally speaking, the terms are equivalent.
+% * <cristina@icmc.usp.br> 2017-08-24T21:43:22.224Z:
+% 
+% > as a knowledge and technological
+% > field  
+% 
+%  ??? não entendi, construção estranha...
+% 
+% ^ <renato.fabbri@gmail.com> 2017-08-29T15:06:18.888Z:
+% 
+% simplifiquei e explicitei um pouco mais, acho que ficou melhor. Ok?
+%
+% ^.
 The main topics of this article are data visualization (or visual analytics)
@@
 \end{Verbatim}
-\noindent to express that there is a 57 years old person called Mary Shastacian which likes reading.
+\noindent to express that there is a 57 year old person called Mary Shastacian who likes reading.
 There are many formats to write/serialize RDF data.
-The example above is written in Turtle and it will be the format used throughout this document.
+The example above is written in Turtle, which will be the format used throughout this document.
 
@@
 \texttt{example:people/mary}.
-One good working example\footnote{Visit \url{http://dbpedia.org/page/Rhesus_macaque}
-and click on the concepts and properties to start browsing the web of linked data.}
-of this is DBPedia~\citep{dbpedia}.
+An interesting working example\footnote{Visit \url{http://dbpedia.org/page/Rhesus_macaque}
+and click on the concepts and properties to start browsing the web of linked data.} is DBPedia~\citep{dbpedia}.
 The process of accessing a URI to find more triples is called \emph{dereferencing the URI} (or simply dereferencing).
@@
 This work is being developed using both approaches.
-The second author is a data visualization specialist which was
+The second author is a data visualization specialist who was
 interviewed by the first author.
-Also, the first author is acquiring a deeper knowledge in the field.
+Also, the first author is engaged in acquiring a deeper knowledge of the field.
 
@@
 and a desirable incorporation of audio to visual analytics given
-audiovisual capabilites of current ordinary computers.
+audiovisual capabilities of current ordinary computers.
 
@@
 Some of the relations bellow are expressed using very recent 
-techniques that are described in~\cite{enhance}.
+techniques described in~\cite{enhance}.
 Their meaning, though, might be easily inferred.
@@
 \begin{center}
-% \includegraphics[height=6.7cm,width=9cm]{../figs/aavo0.01_minimum.png}
-\includegraphics[width=\textwidth]{../figs/aavo0.01_minimum.png}
+% \includegraphics[height=6.7cm,width=9cm]{../figs/aavo0.01_minimum}
+\includegraphics[width=\textwidth]{../figs/aavo0.01_minimum}
 	\caption{AAVO core, discussed in Section~\ref{sec:core}.
@@
 		or 3) presented by means of a Visualization.
-	\item Analysis: a set of procedures used to obtain understanding about
+	\item Analysis: a set of procedures used to gain understanding about
 		Data or a phenomenon.
@@
 \end{itemize}
-\noindent These concepts were not included in AAVO core (e.g. Figure~\ref{fig:minimum})
-because we still deciding on the best way to do so.
-From the definitions above, raises the question:
+\noindent These concepts have not yet been included in AAVO core (e.g. Figure~\ref{fig:minimum})
+because we are still considering the best way to do so.
+From the definitions above,  the question arises:
 should we also include Phenomenon among these core concepts?
@@
 	\item Visualization ``number of dimensions'' real (not double as stated for now).
+% * <cristina@icmc.usp.br> 2017-08-24T21:56:13.120Z:
+% 
+% \item Visualization ``number of dimensions'' real (not double as stated for now).  ??? don´t get it...
+% 
+% 
+% ^ <renato.fabbri@gmail.com> 2017-08-25T20:00:00.908Z:
+% 
+% Double has a certain resolution within the real numbers. A "number of dimensions" need not to be an integer, but might be a real value, not an approximation.
+%
+% ^.
 \end{itemize}
@@
 \begin{center}
-\includegraphics[width=\textwidth]{../figs/aavo0.01_exemplified.png}
+\includegraphics[width=\textwidth]{../figs/aavo0.01_exemplified}
     \caption{Example incorporation of less general concepts to AAVO: Statistical Test, MDS, Timeline, Z-Score, Network, etc.
@@
 There are other uses for AAVO, uncovered in Section~\ref{sec:intro},
-for which there are conceptual models available~\citep{munzner,ward}.
+for which conceptual models are available~\citep{munzner,ward}.
 
@@
 		to other ontologies that are sound, used and maintained.
-	\item Using AAVO for obtaining interesting relations by means of automated
-		inference and for assisting a (audio)visual analytics software.
+	\item Using AAVO to obtain interesting relations by means of automated
+		inference and to assist a (audio)visual analytics software.
 \end{itemize}
@@
 \subsection*{\textit{Acknowledgements}}
-The authors thank CNPq and FAPESP for the funding received while researching the topic of this article,
-the researchers of IFSC/USP and ICMC/USP for the recurrent collaboration in every situation
+The authors thank CNPq and FAPESP for the funding received while researching the topic of this article, the researchers of IFSC/USP and ICMC/USP for the recurrent collaboration in every situation
 where we needed directions for investigation.
@@
 
+\bibitem[Wikipedia, 2017]{wikiVA}
+Visual analytics. (2017, July 1). In Wikipedia, The Free Encyclopedia. Retrieved
+14:51, August 29, 2017
+, from \url{https://en.wikipedia.org/w/index.php?title=Visual_analytics&oldid=788453746}
+
+\bibitem[Ellis \& Mansmann, 2010]{visMaster}
+Ellis, G., \& Mansmann, F. (2010). Mastering the information age solving problems with visual analytics. In Eurographics (Vol. 2, p. 5).
 \end{thebibliography}

### 2017-09-13 latex/article.tex
incorporadas contribs da Cristina
adicionado numero do proj da fapesp

@@
 %%-----------------------------------------------EDIT-----------------------------------------------
-\title{Audiovisual Analytics Vocabulary and Ontology (AAVO): initial core and example expansion}
+\title{AUDIOVISUAL ANALYTICS VOCABULARY AND ONTOLOGY (AAVO): INITIAL CORE AND EXAMPLE EXPANSION}
 
@@
 \subsection*{\textit{Acknowledgements}}
-The authors thank CNPq and FAPESP for the funding received while researching the topic of this article, the researchers of IFSC/USP and ICMC/USP for the recurrent collaboration in every situation
+The authors thank FAPESP (project 2017/05838-3) for the funding received while researching the topic of this article, the researchers of IFSC/USP and ICMC/USP for the recurrent collaboration in every situation
 where we needed directions for investigation.

### 2017-11-06 latex/nuvem/nuvem-template_v1.tex
started article for nuvem

\documentclass[letterpaper,10pt]{article}
\usepackage{nuvem}

\begin{document}

\title{Analytic consideration of }

\author{Marilia M. Pisani$^1$\\
Renato Fabbri$^2$}
\address{$^1$CCNH/UFABC, $^2$IFSC/USP}
\email{$^1$marilia.m.pisani@gmail.com, $^2$renato.fabbri@gmail.com}

\begin{abstract}
Social participation had been growing on the last decade
and had a drawback with the recent rise of conservative political parties 
in various locations of the world.
We aim to present a set of developments on linked social data
and analysis over the last years, with a focus
on the potential use for the civil society and scientific academy.
Various conceptualizations have been gathered by interviewing specialists
and State authorities and have been validated by them and their institutions.
Data from known social networks, such as Facebook, Twitter, IRC and Email,
and from more specialized social participation platforms, have been
	translated to RDF and linked to these conceptualizations in the form of OWL ontologies~\cite{losd,pnud5}.
Social networks were reported as very stable and their language varies
	with connectivity~\cite{tese,stab}.
	Also, resources recommendation and experiments have been performed with such data.
	How are we to articulate the gathering and analysis 
	of such data with the needs of the civil society and the academy?
	Will it sufice to deal with the private and State interests
	that shape our society?
	Does Anthropological Physics yield reasoable strategies
	to collect and analyze data from our society?
\end{abstract}

\section{Introduction}
Social analytics by Schmidt, with its practice of reporting tendencies of the times.
But in contrast, it seems reasonable to put emphasys on issues and analytical outcomes
that are very controversial.

Analítica social. sociedade pela sociedade 
  - Consideração da sociedade pela sociedade:
    * O que fizemos até aqui
    * Qual a questão e contexto atual: com as tecnologias,
    é importante a interface com o ferramental para análise dos dados,
    até para fazer frente aos outros poderes.
    * Como podemos trabalhar o tema: conexão com Nuvem, com Nexos e com IFSC/ICMC.

The appearance of this document reflects the requirements of the style guide.  Since there is no typesetting or copy-editing of summaries, the use of this style guide is critical to provide a consistent appearance.

The first line of the first paragraph of a section or subsection should start flush left. The first line of subsequent paragraphs within the section or subsection should be indented 0.2 in. (0.62 cm).

Use 8.5 in. x 11 in. paper (21.505 cm x 27.83 cm) with 1 in. margins (2.54 cm) on all sides, use 10-point Times New Roman or Palatino font, and do not use hyphens at the end of a line.

\section{What has been done}

\subsection{Social participation data}
\subsection{Ontologies}
\subsection{Critical theory}
\subsection{Fundamental cycle}

\section{What shall we do?}
% keep the data online
% develop an audiovisual analytics software
% maintain partnerships between humanities and hard sciences

\section{Conclusions}

\begin{thebibliography}{99}

\bibitem{krishnan00} E. Krishnan, A. M. Shan, T. Rishi, L. A. Ajith, C. V.
Radhakrishnan, \textit{On-line Tutorial on \LaTeX{}},
``Mathematics'' (Indian \TeX{} Users Group, 2000), \\
\url{http://www.tug.org/tutorials/tugindia/chap11-scr.pdf}.

\bibitem{vantrigt97} C. van Trigt, ``Visual system-response functions and estimating reflectance,''
J. Opt. Soc. Am. A \textbf{14}, 741--755 (1997).

\bibitem{masters93} T. Masters, \emph{Practical Neural Network Recipes in C++} (Academic, 1993).

\bibitem{shoop97} B. L. Shoop, A. H. Sayles, and D. M. Litynski, ``New devices for optoelectronics: smart pixels,''
in \emph{Handbook of Fiber Optic Data Communications},
C. DeCusatis, D. Clement, E. Maass, and R. Lasky, eds. (Academic, 1997), pp. 705--758.

\bibitem{kalman76} R. E. Kalman,``Algebraic aspects of the generalized inverse of a rectangular matrix,'' in
\emph{Proceedings of Advanced Seminar on Generalized Inverse and Applications}, M. Z. Nashed, ed. (Academic, 1976), pp. 111--124.

\bibitem{craig96} R. Craig and B. Gignac, ``High-power 980-nm pump lasers,''
in \emph{Optical Fiber Communication Conference}, Vol. 2 of 1996 OSA Technical Digest Series (Optical Society of America, 1996), paper ThG1.

\bibitem{steup96} D. Steup and J. Weinzierl, ``Resonant THz-meshes,''
presented at the Fourth International Workshop on THz Electronics, Erlangen-Tennenlohe, Germany, 5--6 Sept. 1996.

\end{thebibliography}

\end{document}

### 2017-11-12 latex/nuvem/nuvem-template_v1.tex
caminhado c texto p nuvem

@@
 \usepackage{nuvem}
+\usepackage[utf8]{inputenc}
 
@@
 
-\title{Analytic consideration of }
+\title{Analytic consideration of the society by itself}
 
@@
 	to collect and analyze data from our society?
+	How does the global semantic web of linked data~\cite{GGG}
+	relate to linked social data?
 \end{abstract}
@@
 \section{Introduction}
-Social analytics by Schmidt, with its practice of reporting tendencies of the times.
-But in contrast, it seems reasonable to put emphasys on issues and analytical outcomes
-that are very controversial.
+Social analytics is a term associated to Lars-Hendrik Schmidt~\cite{cite}
+in philosophy.
+One might also relate social analytics to social data mining,
+by extension of the terms: visual analytics, data analytics, social media analytics etc,
+being used nowadays~\cite{wVA,wDA,wSMA,wSNA}.
+Interestingly, both social analytics 
+have among core practices the reporting of tendencies of the times.
+And they put emphasis on issues and analytical outcomes
+that are very controversial and result from recent technological and scientific developments.
 
-Analítica social. sociedade pela sociedade 
-  - Consideração da sociedade pela sociedade:
-    * O que fizemos até aqui
-    * Qual a questão e contexto atual: com as tecnologias,
-    é importante a interface com o ferramental para análise dos dados,
-    até para fazer frente aos outros poderes.
-    * Como podemos trabalhar o tema: conexão com Nuvem, com Nexos e com IFSC/ICMC.
+On the critical theory tradition,
+Marcuse have a central role on
+questioning the conditions
+imposed for living by technological advances~\cite{m1,m2,m3}.
+And the consideration of the authoritarian personality traces~\cite{au}
+have gained relevance due to unforeseen social phenomena
+such as the recent and worldwide rise of the right-winged manifestations
+and political representation~\cite{rise}.
 
-The appearance of this document reflects the requirements of the style guide.
-Since there is no typesetting or copy-editing of summaries, the use of this style guide is critical to provide a consistent appearance.
+On physics and computer sciences,
+both the advent of the complex networks
+theory and the big data practices
+had an impact on the utility of considering
+social systems.
+One enables forecast of properties
+and the use of natural laws for deriving understandings.
+The other provides the data that represents the real systems
+in scales that make them useful for management and
+scientific studies (including experiments as in Section~\ref{exp}).
+Social networking platforms are massively and constantly used and thus fed.
+They are most often elaborate suites of software with
+an interface for the user to interact with other users
+and often makes use of these recent scientific and technological
+developments.
 
-The first line of the first paragraph of a section or subsection should start flush left. The first line of subsequent paragraphs within the section or subsection should be indented 0.2 in. (0.62 cm).
-
-Use 8.5 in. x 11 in. paper (21.505 cm x 27.83 cm) with 1 in. margins (2.54 cm) on all sides, use 10-point Times New Roman or Palatino font, and do not use hyphens at the end of a line.
+In this context,
+we advocate the analytic consideration
+of the Society by itself.
+There are many ways to achieve this,
+and it is performed to some extent.
+Our focus is on the mining of the data
+derived from social platforms.
+It is not very trivial in our context
+of monetary obsession to achieve engagement
+in such analytical undertake.
+Even if there is money making in the process,
+the citizens already have their livelihood activities.
+Therefore, some effort has been given to
+enable this kind of `social participation'.
+Core goals are:
+\begin{itemize}
+	\item Develop the critical view, on specialists and layman,
+		of social media and
+		management performed by State and private sector.
+	\item Make advances in the data available for consideration
+		and the algorithms for analysis.
+	\item Implement or study the possibility implementation
+		of software systems.
+\end{itemize}
 
 \section{What has been done}
+On the following sections, we will focus on
+developments related to social systems and technology.
+There are methods, software and scientific discoveries
+that underpins them~\cite{stab,tese,kolm}.
 
 \subsection{Social participation data}
+Data from Facebook, Twitter, Email lists, IRC Channels, ParticipaBR,
+Cidade Democrática and AA have been translated into linked data (RDF)
+and has been named the LOD (Linked Open Social Data~\cite{lod}).
+Initial OWL ontologies were derived from the data by automated routines.
+The dataset is oriented towards scientific research
+and the development of analytic interfaces~\cite{p1,p2,p3,p4,p5}.
+Even so, it has been used by participants of a SESC oficina~\cite{sesc},
+and has received some attention from the scientific community~\cite{nexos,tese}.
+
 \subsection{Ontologies}
-\subsection{Critical theory}
-\subsection{Fundamental cycle}
+Beyond the ontologies mentioned in the last section,
+there are the social participation ontologies that describe
+social participation instances and mechanisms~\cite{p5}.
+They are: 
+These ontologies are contextualized by the whole semantic web
+(or Giant Global Graph~\cite{ggg})
+and by other social participation ontologies:
+the OPa (Ontology of ParticipaBR~\cite{p5}), the OPS (Ontology of Social Participation~\cite{ops}),
+the OCD (Ontology of Cidade Democrática~\cite{p5}), the OAA (Ontology of AA~\cite{oaa}).
+There are minor ontologies, such as the one about the Caixa Mágica~\cite{cm}.
+These OWL ontologies (and SKOS vocabularies that most of them have) enable
+sound conceptual discussions, navigation by humans and machine (data discovery),
+ automated inference, and data organization, integration and linkage by means of conceptualizations.
+It seems reasonable to use the ontologies to link the social participation data
+and enable a scalable participatory legacy of machine- and human- friendly representation
+of data for navigation, query and analysis, automated inference
+and conceptual considerations.
+
+\subsection{Critical theory and anthropological physics}
+The potential for prejudice and the study of the self
+were considered in~\cite{imp,stab,tese}.
+In summary, the hub, intermediary and peripheral sectorialization
+of a network might be done in a social network.
+It yields a classification of the participants
+by the comparison of the degree (or strength) distribution
+against that of an Erdös-Rényi network.
+Hubs are valued by literature
+(although the intermediary are also reported as authorities
+and structuraly important agents)~\cite{ega},
+which favors the potential for prejudice.
+At the same time, the roles participant
+
+The cold-blooded reception of the reality
+enabled technologial devices such as drones~\cite{drones}.
+
+<últimos trabalhos marilia, e.g. impulso>
+
+The anthropological aspect of observing natural
+phenomena in the social structures
+of the observer him/herself has been
+considered in~\cite{an,an2,imp,tese}.
+Comte talked about anthropological physics,
+but it had the sense which physical (or biological or natural)
+anthropology has, which is a main field of anthropology
+(the other is social anthropology).
+Anthropological physics, in the sense we use and understand correct,
+is physics (research on the natural phenomena of complex networks)
+with anthropological aspects.
+The writing and study of diaries is a traditional ethnographic asset,
+which resembles the collection and study of one's data by the person itself,
+social data.
+The current guidelines are centered on keeping the process open,
+using free and open source tools and file formats,
+and publicly accessible repositories of software, gadgets, data, etc;
+while gathering and studying the social data related to the researcher
+(and other data if needed and afterwards).
+
+\subsection{Self-transparency, AA and the fundamental cycle}
+A system for sharing work, development and research processes
+(actually, any dedication)
+is described in~\cite{aa,aa2}.
+It is the Algorithmic Autoregulation (AA) methodology and it has
+received a number of software implementations.
+It is based on voluntary logging of messages (called shouts)
+about what is being done (reading, coding, etc) that might
+be linked to sessions and screencasts or be blind-reviewed
+by other users.
+The methodology and software have initiated discussions
+e.g. about documenting academic dedication not only though
+documents (e.g. articles, books) and participation (e.g. conferences, thesis defense),
+which would enable one to dedicate more time to individual
+tasks and make more paced and relevant contributions.
+
+Also, the conception of the
+
+\subsection{Audiovisual Analytics platform}
+for social data, with focus on network and textual
+		data.
 
 \section{What shall we do?}
+Most certainly we will
+make the LOSD and social participation ontologies (Sections~\ref{losd} and~\cite{ont})
+available e.g. on DataHub and Data.World.
+In the Nexos research network, we
+considered a number of times to study authoritarian personality traces~\cite{au}
+on this data.
+And to further consider the Anthropological Physics aspects that
+arise from studying our own social structures~\cite{af,af2}.
+The self-transparency is having some attention in scientific writings
+and hacks~\cite{aa,aa2} but seems to be lacking user bases.
+The developments for Audiovisual Analytics are incipient
+and has the development focus of only one researcher and developer.
+The Nexos research network is very active nation wide.
+It has an overall tendency to tackle subjects within the
+critical theory tradition and thus considers the society
+and potential enhancements through criticism of its organization.
+
+Research groups @Nuvem of Cloud Computing and Intelligent Societies have 
+further context to share about these issues.
+For example, are there SparQL endpoints with social participation
+(including self-transparency) data?
+Is it necessary to contact lawyers to
+better know the limits of our possibilities
+to gather and research our social data within
+the Anthropological Physics perspective
+and are there very well known guidelines?
+Is there more data participatory linked data available
+in Brazil? Ontologies?
+We might use directions 
+on better linking LOSD to the semantic web
+(DBPedia, other participatory data, etc),
+and for a reasonable way to keep the data online
+(through DataHub or Data.World or both?
+A Pubby-like interface?),
+and to develop an audiovisual analytics
+software
+(persistence, web?, analysis methods, audiovisual rendering).
+
+How to enable a self-transparency user base?
+Where to keep the linked data?
+How to manage the ontologies and keep their
+constant development (as needed and predicted in the literature)?
+How to achieve a reasonable posture
+and use of our social (open linked) data?
+Is it possible to have an audiovisual analytics
+platform with our own (scraped) social data,
+with facilities for media rendering,
+interaction experiments (collection and diffusion of information),
+that enables the user base and interested parties to
+collective gather social data, analysis and conceptualizations?
+Is the societal consideration of our social data
+relevant for equilibrium with the State and private sectors?
+
 % keep the data online
@@
 % maintain partnerships between humanities and hard sciences
+
+    * Como podemos trabalhar o tema:
+    conexão com Nuvem, com Nexos e com IFSC/ICMC.
 

### 2017-11-12 latex/nuvem/nuvem-template_v1.tex
terminada primeira versao

--- 2017-11-06 latex/nuvem/nuvem-template_v1.tex
@@
 \usepackage{nuvem}
+\usepackage[usenames,dvipsnames]{xcolor}
+\usepackage{hyperref} 	%% Use to fix Figure or Table: ex: \begin{table}[H]
+\hypersetup{
+	%pagebackref=true,
+	pdfcreator={LaTeX with abnTeX2},
+	pdfkeywords={abnt}{latex}{abntex}{USPSC}{trabalho acadêmico}, 
+	colorlinks=true,       		% false: boxed links; true: colored links
+	linkcolor=blue,          	% color of internal links
+	citecolor=blue,        		% color of links to bibliography
+	filecolor=magenta,      		% color of file links
+	urlcolor=blue,
+	allbordercolors=black,
+	bookmarksdepth=4
+}
+\usepackage[utf8]{inputenc}
 
@@
 \begin{abstract}
-Social participation had been growing on the last decade
-and had a drawback with the recent rise of conservative political parties 
+Social participation has been growing on the last decade though with a considerable
+drawback with the recent rise of conservative political parties 
 in various locations of the world.
-We aim to present a set of developments on linked social data
-and analysis over the last years, with a focus
+We aim to consider a set of developments on linked social data
+and analysis performed over the last years, with a focus
 on the potential use for the civil society and scientific academy.
@@
 	of such data with the needs of the civil society and the academy?
-	Will it sufice to deal with the private and State interests
+	Will it be valuable to deal with the private and State interests
 	that shape our society?
-	Does Anthropological Physics yield reasoable strategies
-	to collect and analyze data from our society?
+	Does Anthropological Physics yield reasonable strategies
+	to collect and analyze data from our social structures?
+	How does the global semantic web of linked data~\cite{losd}
+	relate to linked social data?
 \end{abstract}
@@
 \section{Introduction}
-Social analytics by Schmidt, with its practice of reporting tendencies of the times.
-But in contrast, it seems reasonable to put emphasys on issues and analytical outcomes
-that are very controversial.
+Social analytics is a term associated to Lars-Hendrik Schmidt~\cite{wSA}
+in philosophy.
+One might also relate social analytics to social data mining
+by extension of the terms: visual analytics, data analytics, social media analytics etc,
+being used nowadays (e.g. look for these terms in the English language Wikipedia).
+Interestingly, both social analytics 
+have among core practices the reporting of tendencies of the times.
+And they put emphasis on issues and analytical outcomes
+that are very controversial and result from recent technological and scientific developments.
 
-Analítica social. sociedade pela sociedade 
-  - Consideração da sociedade pela sociedade:
-    * O que fizemos até aqui
-    * Qual a questão e contexto atual: com as tecnologias,
-    é importante a interface com o ferramental para análise dos dados,
-    até para fazer frente aos outros poderes.
-    * Como podemos trabalhar o tema: conexão com Nuvem, com Nexos e com IFSC/ICMC.
+On the critical theory tradition,
+Marcuse have a central role on
+questioning the conditions
+imposed for living by technological advances~\cite{m1,m2,m3}.
+And the consideration of the authoritarian personality traces~\cite{au}
+have gained relevance due to unforeseen social phenomena
+such as the recent and worldwide rise of the right-winged manifestations
+and political representation~\cite{rise,rise2}.
 
-The appearance of this document reflects the requirements of the style guide.
-Since there is no typesetting or copy-editing of summaries, the use of this style guide is critical to provide a consistent appearance.
+On physics and computer sciences,
+both the advent of the complex networks
+theory and the big data practices
+had an impact on the utility of considering
+social systems.
+One enables forecast of properties
+and the use of natural laws for deriving understandings.
+The other provides the data that represents the real systems
+in scales that make them useful for management and
+scientific studies (including experiments as in Sections~\ref{exp} and~\ref{self}).
+Social networking platforms are massively and constantly used and thus fed.
+They are most often elaborate suites of software with
+an interface for the user to interact with other users
+and often makes use of these recent scientific and technological
+developments.\cite{tese}
 
-The first line of the first paragraph of a section or subsection should start flush left. The first line of subsequent paragraphs within the section or subsection should be indented 0.2 in. (0.62 cm).
-
-Use 8.5 in. x 11 in. paper (21.505 cm x 27.83 cm) with 1 in. margins (2.54 cm) on all sides, use 10-point Times New Roman or Palatino font, and do not use hyphens at the end of a line.
+In this context,
+we advocate the analytic consideration
+of the Society by itself.
+There are many ways to achieve this,
+and it is performed to some extent~\cite{scriptLattes,losd,p5,an}.
+Our focus here is on the mining of the data
+derived from social platforms.
+It is not very trivial in our context
+of monetary obsession to achieve engagement
+in such analytical undertake.
+Even if there is money making in the process,
+the citizens already have their livelihood activities.
+Therefore, some effort has been given to
+enable this kind of `social participation'.
+Core goals are:
+\begin{itemize}
+	\item Develop the critical view, on specialists and layman,
+		of social media and
+		management performed by the State and the private sector.
+	\item Make advances in the data available for consideration
+		and the algorithms for analysis.
+	\item Implement or study the possibility of implementation
+		of software systems.
+\end{itemize}
 
 \section{What has been done}
+On the following sections, we will focus on
+developments related to social systems and technology.
+There are methods, software and scientific discoveries
+that underpins them~\cite{stab,tese,kolm}.
 
-\subsection{Social participation data}
-\subsection{Ontologies}
-\subsection{Critical theory}
-\subsection{Fundamental cycle}
+\subsection{Social participation data}\label{losd}
+Data from Facebook, Twitter, Email lists, IRC Channels, ParticipaBR,
+Cidade Democrática and AA have been translated into linked data (RDF)
+and has been named the LOD (Linked Open Social Data~\cite{losd}).
+Initial OWL ontologies were derived from the data by automated routines.
+The dataset is oriented towards general scientific research
+and the development of analytic interfaces~\cite{p1,p2,p3,p4,p5}.
+Even so, it has been used e.g. by participants of a SESC workshop
+and has received some attention from the scientific community~\cite{mariliaNexos,tese},
+and mining of the data revealed very stable patterns as predicted by natural laws
+and strong differentiation of the language of participants across hubs,
+intermediary and the periphery~\cite{stab,tese}.
+
+\subsection{Ontologies}\label{ont}
+Beyond the ontologies mentioned in the last section,
+there are social participation ontologies that describe
+social participation instances and mechanisms~\cite{p5}.
+Such instances and mechanisms are:
+Conference (Conferência), Forum (Fórum),
+Committee (Comitê), Council (Conselho), Ombudsman (Ouvidoria),
+Public Consultation (Consulta Pública), Dialog Table (Mesa de diálogo),
+Monitoring Table (Mesa de Monitoramento), Intercouncil Forum (Fóruns Interconselhos),
+Audience (Audiëncia),
+Virtual Environment (Ambiênte Virtual).
+They received additional ontologies relating them to documents and
+larger scopes.
+
+These ontologies are contextualized by the whole semantic web
+(or Giant Global Graph~\cite{losd})
+and by other social participation ontologies:
+the OPa (Ontology of ParticipaBR~\cite{p5}), the OPS (Ontology of Social Participation~\cite{ops}),
+the OCD (Ontology of Cidade Democrática~\cite{p5}), the OAA (Ontology of AA~\cite{p5}).
+There are minor ontologies, such as the one about the Magic Box (Caixa Mágica~\cite{cm},
+a social participation wifi gadget).
+These OWL ontologies (and the SKOS vocabularies which most of them have, following
+a pattern suggested in~\cite{ont}) enable
+sound conceptual discussions, navigation by humans and machine (data discovery),
+ automated inference; and data organization, integration and linkage by means of conceptualizations.
+It seems reasonable to use the ontologies to link the social participation data
+and enable a scalable participatory legacy of machine- and human- friendly representation
+of data for navigation, query and analysis, automated inference
+and conceptual considerations.
+
+\subsection{Critical theory and anthropological physics}\label{exp}
+The potential for prejudice and the study of the self
+were considered in~\cite{imp,stab,tese}.
+In summary, the hub, intermediary and peripheral sectorialization
+of a network might be done in a social network.
+It yields a classification of the participants
+by the comparison of the degree (or strength) distribution
+against that of an Erdös-Rényi network.
+Hubs are valued by literature
+(although the intermediary are also reported as authorities
+and structuraly most important agents)~\cite{ega},
+which favors the potential for prejudice
+of a quantitative classification (or typology)
+of human beings..
+At the same time, the roles participant
+vary in time constantly~\cite{stab,tese}
+and across scales and networks,
+which minimized the potential
+for prejudice of such classification.
+
+We also considered the cold-blooded reception of the reality
+enabled technologial devices such as drones~\cite{drones}.
+(((últimos trabalhos marilia, e.g. impulso,
+talvez com foco em tecnologia, talvez passar
+este parágrafo para o último e fechar a seção com ele))))
+
+The Anthropological Physics is concerned
+with the observation of natural
+phenomena in the social structures
+of the observer him/herself and has been
+considered in.
+Comte talked about anthropological physics,
+but it had the sense which physical (or biological or natural)
+anthropology has, which is a branch of anthropology
+(the other is social anthropology).
+Anthropological physics, in the sense we use and understand correct,
+is physics (research on the natural phenomena mainly of complex networks)
+with anthropological aspects.
+The writing and study of diaries is a traditional ethnographic asset,
+which resembles the collection and study of one's data by the person itself,
+social data.
+The current guidelines are centered on keeping the process open,
+using free and open source tools and file formats,
+and publicly accessible repositories of software, gadgets, data, writings, etc;
+while gathering and studying the social data related to the researcher
+(and other data if needed and afterwards).~\cite{an,an2,imp,tese}
+
+\subsection{Self-transparency, AA and the fundamental cycle}\label{self}
+A system for sharing work, development and research processes
+(actually, any dedication)
+is described in~\cite{aa,aa2}.
+It is the Algorithmic Autoregulation (AA) methodology and it has
+received a number of software implementations.
+It is based on voluntary logging of messages (called shouts)
+about what is being done (reading, coding, etc) that might
+be linked to sessions and screencasts or be blind-reviewed
+by other users.
+The methodology and software have initiated discussions
+e.g. about documenting academic dedication not only though
+traditional documents (e.g. articles, books) and participation (e.g. conferences, thesis defense),
+and currently performed e.g. by Lattes.
+Such documentation (log of dedications)
+would enable one to dedicate more time to individual
+tasks and make more paced and relevant contributions.
+
+The concept has reached other initiatives, such
+as the Brazilian federal portal of social participation
+ParticipaBR.
+And have incited discussions that yield interesting concepts,
+such as the fundamental cycle,
+which is an idealized interaction between
+Society and a management sector of it,
+such as the State.
+The individuals that feel the need to
+change the way living is imposed by the State
+should be able to act on the State,
+have results and get paid for the time they put
+into performing the social participation.
+The existence of such a cycle would erradicate
+extreme poverty, as the poor would be paid
+to act and change State measures for the poor for
+or a community.
+Thus, the ideal fundamental cycle is a social
+participation model which is very shabbily practiced
+and is defined by giving the one in need good conditions
+to change the context which puts he/she in need.
+The mechanisms or institutions that give the
+individual the conditions and means to act
+is then called the State or Government or,
+more accurately nowadays, the Institutionalized Power.
+Notice that, as such, the institutionalized power
+have the goal to allow the interference 
+of the individual in its management
+and that this is extended to the private initiatives.
+Social participation in the private sector
+seems to be very incipient.
+
+\subsection{Audiovisual Analytics platform}
+Described in~\cite{nuvem2}, it is an envisioned audiovisual analytics
+software for social data, with focus on network and textual data.
+It will also be sent to the I Workshop @NUVEM, together with
+this document, for separate appreciation.
 
 \section{What shall we do?}
-% keep the data online
-% develop an audiovisual analytics software
-% maintain partnerships between humanities and hard sciences
+Most certainly we will
+make the LOSD and social participation ontologies (Sections~\ref{losd} and~\ref{ont})
+available e.g. on DataHub and Data.World.
+In the Nexos research network, we
+considered a number of times to study authoritarian personality traces~\cite{au}
+on this data.
+And to further consider the Anthropological Physics aspects that
+arise from studying our own social structures~\cite{an,an2}.
+The self-transparency is having some attention in scientific writings
+and hacks~\cite{aa,aa2} but seems to be lacking user bases.
+The developments for Audiovisual Analytics are incipient
+and has the development focus of only one researcher and developer.
+The Nexos research network is very active nation wide.
+It has an overall tendency to tackle subjects within the
+critical theory tradition and thus considers the society
+and potential enhancements through criticism of its organization,
+which is valuable for the technological developments being made.
 
-\section{Conclusions}
+Research groups @Nuvem of Cloud Computing and Intelligent Societies might have 
+further context to share about these issues.
+For example, are there SparQL endpoints with (Brazilian) social participation
+(including self-transparency) data?
+Is it necessary to contact lawyers to
+better know the limits of our possibilities
+to gather and research our social data within
+the Anthropological Physics perspective
+and are there well known guidelines?
+Is there more participatory linked data available
+in Brazil? Ontologies?
+We might use directions 
+on better linking LOSD to the semantic web
+(DBPedia, other participatory data, etc),
+and for a reasonable way to keep the data online
+(through DataHub or Data.World or both?
+A Pubby-like interface?),
+and to develop an audiovisual analytics
+software
+(persistence, web?, analysis methods, audiovisual rendering).
+
+Other questions we have that might benefit from discussion and entail collaboration are:
+how to enable a self-transparency user base?
+Where to keep the linked data?
+How to manage the ontologies and keep their
+constant development (as needed and predicted in the literature)?
+How to achieve a reasonable 
+use of our social (open linked) data?
+Is it possible to have an audiovisual analytics
+platform with our own (scraped) social data,
+with facilities for media rendering and
+interaction experiments (collection and diffusion of information),
+that enables the user base and interested parties to
+collective gather social data, analysis and conceptualizations?
+Is the societal consideration of our social data
+relevant for equilibrium with the State and private sectors?
+Is it possible to achieve social participation in the private sector,
+e.g. to have civil representation in the chair of companies,
+to regulate matters such as YouTube adds in Google,
+nutrient standards in McDonald's,
+and other things in other major and minor institutions?
 
@@
 
-\bibitem{krishnan00} E. Krishnan, A. M. Shan, T. Rishi, L. A. Ajith, C. V.
-Radhakrishnan, \textit{On-line Tutorial on \LaTeX{}},
-``Mathematics'' (Indian \TeX{} Users Group, 2000), \\
-\url{http://www.tug.org/tutorials/tugindia/chap11-scr.pdf}.
+\bibitem{losd}
+	Fabbri, R., \& de Oliveira, O. N. (2016). Linked Open Social Database. Github repositories, from \url{https://github.com/ttm/linkedOpenSocialDatabase/raw/master/paper.pdf}
 
-\bibitem{vantrigt97} C. van Trigt, ``Visual system-response functions and estimating reflectance,''
-J. Opt. Soc. Am. A \textbf{14}, 741--755 (1997).
+\bibitem{pnud5}
+Fabbri, R. (2014). Social participation ontologies and rules to fuel a social participation linked data cloud.
+	Technical report for the United Nations Development Program.
+		From \url{https://github.com/ttm/pnud5/raw/master/latex/produto.pdf}
 
-\bibitem{masters93} T. Masters, \emph{Practical Neural Network Recipes in C++} (Academic, 1993).
+\bibitem{tese}
+Fabbri, R. (2017). Topological stability and textual differentiation in human interaction networks:
+		statistical analysis, visualization and linked data. Doctoral thesis.
+		From \url{https://github.com/ttm/thesis/raw/master/thesis-rfabbri.pdf}
 
-\bibitem{shoop97} B. L. Shoop, A. H. Sayles, and D. M. Litynski, ``New devices for optoelectronics: smart pixels,''
-in \emph{Handbook of Fiber Optic Data Communications},
-C. DeCusatis, D. Clement, E. Maass, and R. Lasky, eds. (Academic, 1997), pp. 705--758.
+\bibitem{stab}
+Fabbri, R., Fabbri, R., Antunes, D. C., Pisani, M. M., \& de Oliveira, O. N. (2017). Temporal stability in human interaction networks. Physica A: Statistical Mechanics and its Applications.
 
-\bibitem{kalman76} R. E. Kalman,``Algebraic aspects of the generalized inverse of a rectangular matrix,'' in
-\emph{Proceedings of Advanced Seminar on Generalized Inverse and Applications}, M. Z. Nashed, ed. (Academic, 1976), pp. 111--124.
+\bibitem{wSA}
+	Social analytics. (2017, February 5). In Wikipedia, The Free Encyclopedia. Retrieved 11:37, November 12, 2017, from \url{https://en.wikipedia.org/w/index.php?title=Social\_analytics\&oldid=763744127}
 
-\bibitem{craig96} R. Craig and B. Gignac, ``High-power 980-nm pump lasers,''
-in \emph{Optical Fiber Communication Conference}, Vol. 2 of 1996 OSA Technical Digest Series (Optical Society of America, 1996), paper ThG1.
+\bibitem{au}
+	Adorno, T. W., Frenkel-Brunswik, E., Levinson, D. J., \& Sanford, R. N. (1950). The authoritarian personality.
 
-\bibitem{steup96} D. Steup and J. Weinzierl, ``Resonant THz-meshes,''
-presented at the Fourth International Workshop on THz Electronics, Erlangen-Tennenlohe, Germany, 5--6 Sept. 1996.
+\bibitem{rise}
+Kaufmann, E. (2017). Interview with Eric Kaufmann: cultural values and the rise of right-wing populism in the West. British Politics and Policy at LSE.
 
+\bibitem{rise2}
+	Gaarsted, J. B., \& Agustín, L. R. (2017). The rise of right-wing populism.
+
+\bibitem{scriptLattes}
+	Mena-Chalco, J.
+P., \& Júnior, C.
+E.
+S.
+A.
+R.
+(2013).
+Prospecção de dados acadêmicos de currículos Lattes através de Scriptlattes.
+Bibliometria e Cientometria: reflexões teóricas e interfaces.
+São Carlos: Pedro \& João.
+
+\bibitem{kolm} 
+  R.~Fabbri and F.~G.~De León,
+  ``A Statistical Distance Derived From The Kolmogorov-Smirnov Test: specification, reference measures (benchmarks) and example uses,''
+  Anais do XX ENMC - Encontro Nacional de Modelagem Computacional e
+  VIII ECTM - Encontro de Ci\^encias e Tecnologia de Materiais, Nova Friburgo,
+  RJ - 16 a 19 Outubro 2017
+		[arXiv:1711.00761 [physics.data-an]]. Available at: \url{https://arxiv.org/abs/1711.00761}.
+
+\bibitem{p1}
+	Fabbri, R. Primeira ontologia do Portal Federal de Participação Social Descrição e código OWL: Ontologia do ParticipaBR, 2014.
+    United Nations Development Programme and Brazilian Presidency of the Republic.
+Available at \url{https://sourceforge.net/p/labmacambira/fimDoMundo/ci/master/tree/textos/ontologia/ontologiaParticipa\_.pdf}
+
+\bibitem{p2}
+    Fabbri, R. United Nations Development Programme: ParticipaBR context, triplification of data and example of usage, 2014.
+    Available at \url{http://sourceforge.net/p/labmacambira/fimDoMundo/ci/master/tree/textos/SparQL/triplificaDisponibiliza.pdf?format=raw}
+
+\bibitem{p3}
+    Fabbri, R. United Nations Development Programme: Tools for content classification in the ParticipaBR Brazilian federal portal of social participation, 2014.
+    Available at \url{https://github.com/ttm/pnud3/blob/master/latex/produto.pdf?raw=true}
+
+\bibitem{p4}
+    Fabbri, R. United Nations Development Programme: Adaptations and increments for the ParticipaBR Brazilian federal portal of social participation, 2014.
+    Available at \url{https://github.com/ttm/pnud4/blob/master/latex/produto.pdf?raw=true}
+
+\bibitem{p5}
+Fabbri, R. Social participation ontologies and rules to fuel a social participation linked data cloud, 2014.
+	Technical report for the United Nations Development Program, made with the Brazilian Presidency.
+		Available at \url{https://github.com/ttm/pnud5/raw/master/latex/produto.pdf}
+
+\bibitem{p5}
+Fabbri, R. Social participation ontologies and rules to fuel a social participation linked data cloud, 2014.
+	Technical report for the United Nations Development Program, made with the Brazilian Presidency.
+		Available at \url{https://github.com/ttm/pnud5/raw/master/latex/produto.pdf}
+
+\bibitem{ops}
+	Fabbri, R., de Luna, R.
+B., Martins, R.
+A.
+P., Amanqui, F.
+K.
+M., Moreira, D.
+D.
+A., Oliveira Jr, O.
+N.
+O.
+(2015).
+Social Participation Ontology: community documentation, enhancements and use examples.
+arXiv preprint arXiv:1501.02662.
+Available at \url{https://github.com/ttm/ops/raw/master/plosOne/plosComImagens.pdf}
+
+\bibitem{cm}
+	Fabbri, R. Magic Box Ontology.
+		Available at \url{https://github.com/ttm/pnud5/raw/master/latex/produto.pdf}
+
+\bibitem{imp}
+	Fabbri, R. Redes complexas para redes sociais: introdução, aspectos críticos e software, 2017.
+		Revista Impulso,
+		ISSN 2236-9767,
+		DOI 10.15600/2236-9767/impulso.v27n69p65-75.
+		Disponível em
+		\url{https://www.metodista.br/revistas/revistas-unimep/index.php/impulso/article/view/3367}.
+
+\bibitem{ont} 
+  R.~Fabbri and M. Pisani,
+  ``Enhancements of linked data expressiveness for ontologies,''
+  Anais do XX ENMC - Encontro Nacional de Modelagem Computacional e
+  VIII ECTM - Encontro de Ci\^encias e Tecnologia de Materiais, Nova Friburgo,
+  RJ - 16 a 19 Outubro 2017
+		Available at: \url{https://github.com/ttm/equanimousScalefree/raw/master/article/article.pdf}.
+
+\bibitem{ega} 
+  R.~Fabbri and M. Pisani,
+  ``Egalitarian aspects of scale-free networks,''
+  Anais do XX ENMC - Encontro Nacional de Modelagem Computacional e
+  VIII ECTM - Encontro de Ci\^encias e Tecnologia de Materiais, Nova Friburgo,
+  RJ - 16 a 19 Outubro 2017
+		Available at: \url{https://github.com/ttm/equanimousScalefree/raw/master/article/article.pdf}.
+
+\bibitem{an}
+	Fabbri, R. What are you and I? [anthropological physics fundamentals], 2015. Available at \url{https://www.academia.edu/10356773/What\_are\_you\_and\_I\_anthropological\_physics\_fundamentals\_}
+
+\bibitem{an2}
+	Anthropological physics and social psychology in the critical research of networks. Complex Networks Digital Campus (CS-DC'15).
+	Em \url{http://cs-dc-15.org/papers/cognition/social-psychology/anthropological-physics-and-social-psychology-in-the-critical-research-of-networks/}
+
+\bibitem{aa}
+	Fabbri, R., Fabbri, R., Vieira, V., Penalva, D., Shiga, D., Mendonça, M., Negrao, A., Zambianchi, L., \& Thumé, G.
+(2013).
+AA: The Algorithmic Autoregulation (Distributed Software Development) Methodology.
+RESI.
+From \url{https://arxiv.org/abs/1604.08255}
+
+\bibitem{aa2} 
+  R.~Fabbri,
+  ``The Algorithmic-Autoregulation (AA) Methodology and Software: a collective focus on self-transparency,''
+  Anais do XX ENMC - Encontro Nacional de Modelagem Computacional e
+  VIII ECTM - Encontro de Ci\^encias e Tecnologia de Materiais, Nova Friburgo,
+  RJ - 16 a 19 Outubro 2017
+		Available at: \url{https://github.com/ttm/ensaaio/raw/master/emc/article.pdf}.
+
+\bibitem{nuvem2} 
+  R.~Fabbri and M. Pisani
+  ``Audiovisual Analytics of social data,''
+  I Workshop @NUVEM, UFABC campus de Santo André,
+		Available at: \url{}.
 \end{thebibliography}

### 2017-11-13 latex/nuvem/nuvem2.tex
first skech of audiovisual analytics

\documentclass[letterpaper,10pt]{article}
\usepackage{nuvem}

\begin{document}

\title{Audiovisual Analytics of social linked data}

\author{Renato Fabbri$^1$\\
	Marilia M. Pisani$^2$
}
\address{$^1$IFSC/USP, $^2$CCNH/UFABC}
\email{$^1$renato.fabbri@gmail.com, $^2$marilia.m.pisani@gmail.com}

\begin{abstract}
	Audiovisual analytics of social data is useful for State and private
	management, scientific and individual endeavors.
	If you have data as information in a computer, and wants to
	transmit it to an observer or user, the use of visual cues is an obvious,
	and sometimes considered optimized, choice because of the efficiency and
	complexity of our visual cognition.
	Our sonic cognition is sometimes considered of greater complexity,
	due to patterns of music and spoken language.
	How to correctly formalize the mapping of data to physical stimuli
	and then to the perceived stimuli
	respecting the Webner-Fechner and Steven's laws?
	Related to sound, most traditional notation and musical qualities
	respect the Webner-Fechner law, with note grids (e.g. semitones, octaves),
	respecting the logarithm of frequency: iterval$=log_2(f_1/f_0)$,
	loudness$=log_10(p_1/p_0)$
	Duration seems to be conceptualized as such
	because musical notation uses the division in two (simple, imperfectus),
	amd three (perfectus) yielding a grid having
	$d_i=d_{i+1}/2$ and $d_i=d_{i+1}/3$.
	$durations=2^{x-i}$ and $=3^{x-i}$.
	Which implies $log_2(dur) = x-i$.
	A Webner-Fechner compliant relation
	between the number of the durations and
	depth of the division x-i = (4, 8, 16) for simple,
	(3, 9, 12) for compound,
	(5,7,13) complex.
	6 is compound in traditional music theory,
	but is the only one that relates periods of 2 and 3,
	subdivision of 2 and 3.
\end{abstract}

\section{Introduction}
Linked data is useful for data integration from different sources,
for semantic representation and manipulation of data,
such as by automated inferences,
and sound conceptual considerations of both domain and context
knowledge.
The mapping of data to audiovisual cues
involve a number of routines, conceptual frameworks,
and data.
The scaling and integration factors are decisive in setting
of Big Data such as in social data mining.
The study of data that expresses human social systems
poses ethical and conceptual challenges~\cite{an,an2}.
An overview of the Linked Open Social Data gathered
in RDF and related ontologies and vocabularies
can be found in~\cite{nuvem1,tese,losd}.
Some limitations of current linked data standards
are described and potential solutions given in~\cite{ont}.

\section{Current Framework}\label{current}
We have strived to obtain relevant algebraic, empiric, conceptual,
analyses of social structures~\cite{tese,stab},
often supported by software, which is a very formal representation
of what is being performed (although also very incomplete).
In this section, we focus on current efforts around Audiovisual Analytics.

\subsection{AAVO}
AAVO is a first tentative ontology for Audiovisual Analytics.
Figure~\cite{aavo} depicts the AAVO Core,
and both Core and Extended AAVO are well described in~\cite{aavoA}.
The following concepts are envisioned for the core:

And it might receive other layers of conceptualizations,
such as for data type, colors, charts, etc.

Its purposes are: automated inferences and recommendations on
contexts involving data visualization;
enable relevant theoretical discussions by having an
established and formal conceptualization;
provide a representation of data that is friendly for
both humans and machines (in browsing, discovery, inference);
underpin data integration;
provide the conceptual and data architecture for
the audiovisual analytics software described in
the next section.

\subsection{An audiovisual analytics platform of social systems}
A software system, a platform for audiovisual analytics of big data
have been envisioned which have the following characteristics:
\begin{itemize}
  \item Interface has capabilities of making automated and periodic changes of
	  the presentation of the content (data and analyses).
		This entails a fractal and musical consideration
		of the audiovisualzations.
		User might pause and set changing patterns..
  \item The simultaneous use of both the vision and hearing channels.
	  User might mute or focus on the sonic cues.
  \item The system has the purpose of providing an aesthetic experience
	  to the user.
		This includes facilities to render audiovisual media
		for aesthetic appreciation or attractive representation of
		analyses~\cite{tufte}.
  \item Keeps records of state through users and sessions.
	  Each session has a set of states, each state a set
		of interface parametrization and data references.
	Annotations might be linked to users, sessions, states, data,
		analysis methods, widgets, etc; and categorized
		by keywords or simple values.
  \item Persistence (probably through the web browser's sandbox)
	  to allow a user to reload already downloaded software
		and data,
		and to allow a user to work offline.
  \item Sharing of sessions, states, data, and annotations among users,
	  preferably in real time (as achieved through Meteor.js).
		Sharing of media to non-users and for presenting results.
  \item A written interface to control the system which should be
	  suitable for scripting, one line commands, keyboard shortcuts.
		(The written interface expresses a very broad space
		of possible commands.)
  \item Linked data to achieve a formal and browse-friendly representation of data for both humans and machines;
	  for a unified consideration of both data and knowledge architecture;
facilitates the linkage of metadata from users;
facilitates resources recommendations.
  \item Emphasis on the analysis of networks and text (and their simultaneous analysis),
	  reason why it is fit for social networks, but also for general complex networks
		and textual data.
  \item Designed to be enhanced with use by means of tests with users, of analyses of sessions and states
	  (maybe have facilities to record mouse movement, clicks and keyboard strokes).
\end{itemize}

The temporal dynamic of the interface is useful to do a rapid scan of the data by parts
Observing data by parts is a core procedure in mining big data.
Persistence is also valued for the purpose of analyzing big data.
Complex networks and text yield a broad user base
and set of knowledge fields which might be studied withing the system.
The focus on social systems entails analysis results of interest
to both the scientific research, and State and private sectors.
The artistic course of current and envisioned developments
is useful for formal documents but also favors the engagement
of layman.

\subsection{Perceptual framework}
Link to AAVO: the three sampling groups.
\subsection{Other: Erdös Sectioning, KS-derived statistics, Wordnet,
text mining, stability and differentiation, partnerships}
\subsection{What to do?}
We will surely integrate the Hypothesis, Goals, Interaction
concepts into the AAVO Core.
We should develop the system in Section~\ref{sys} to some extent,
maybe for concept proofs or simple prototyping.

With the research groups @Nuvem, 
relate the missing concepts in AAVO,
enhance the software design of the system,
Know about the right tools for visualizing
ontologies and social data.
For knowing about pertinent analysis methods
and known conceptualizations.
Virtual reality, visualization and societal groups
might want to formalize their own conceptualizations,
and concepts related to AAVO might have similar developments
or which are confluent (e.g. symbiotic)
with Section~\ref{current}: software
theoretical framework, research and institutional experiments and equipment,
linked data, and other initiatives or knowledge.

\begin{thebibliography}{99}

\bibitem{krishnan00} E. Krishnan, A. M. Shan, T. Rishi, L. A. Ajith, C. V.
Radhakrishnan, \textit{On-line Tutorial on \LaTeX{}},
``Mathematics'' (Indian \TeX{} Users Group, 2000), \\
\url{http://www.tug.org/tutorials/tugindia/chap11-scr.pdf}.

\bibitem{vantrigt97} C. van Trigt, ``Visual system-response functions and estimating reflectance,''
J. Opt. Soc. Am. A \textbf{14}, 741--755 (1997).

\bibitem{masters93} T. Masters, \emph{Practical Neural Network Recipes in C++} (Academic, 1993).

\bibitem{shoop97} B. L. Shoop, A. H. Sayles, and D. M. Litynski, ``New devices for optoelectronics: smart pixels,''
in \emph{Handbook of Fiber Optic Data Communications},
C. DeCusatis, D. Clement, E. Maass, and R. Lasky, eds. (Academic, 1997), pp. 705--758.

\bibitem{kalman76} R. E. Kalman,``Algebraic aspects of the generalized inverse of a rectangular matrix,'' in
\emph{Proceedings of Advanced Seminar on Generalized Inverse and Applications}, M. Z. Nashed, ed. (Academic, 1976), pp. 111--124.

\bibitem{craig96} R. Craig and B. Gignac, ``High-power 980-nm pump lasers,''
in \emph{Optical Fiber Communication Conference}, Vol. 2 of 1996 OSA Technical Digest Series (Optical Society of America, 1996), paper ThG1.

\bibitem{steup96} D. Steup and J. Weinzierl, ``Resonant THz-meshes,''
presented at the Fourth International Workshop on THz Electronics, Erlangen-Tennenlohe, Germany, 5--6 Sept. 1996.

\end{thebibliography}

\end{document}

### 2017-11-13 latex/nuvem/nuvem1.tex
adicionada primeira versao do artigo p nuvem
incorporada contribs da marilia no nuvem1

--- 2017-11-12 latex/nuvem/nuvem-template_v1.tex
@@
 \usepackage[usenames,dvipsnames]{xcolor}
-\usepackage{hyperref} 	%% Use to fix Figure or Table: ex: \begin{table}[H]
+\usepackage{hyperref}
 \hypersetup{
@@
 questioning the conditions
-imposed for living by technological advances~\cite{m1,m2,m3}.
+imposed for living by technological advances~\cite{m1,m2,m3,m4}.
 And the consideration of the authoritarian personality traces~\cite{au}
@@
 
+Violence, prejudice and authoritarianism,
+and the authoritarian personality and technology,
+are frequently co-occurring themes in the two dossier of the
+Nexos Networks for Critical Theory and Interdisciplinary Research~\cite{nexos1}.
 We also considered the cold-blooded reception of the reality
-enabled technologial devices such as drones~\cite{drones}.
-(((últimos trabalhos marilia, e.g. impulso,
-talvez com foco em tecnologia, talvez passar
-este parágrafo para o último e fechar a seção com ele))))
+enabled technological devices such as drones~\cite{drones}.
 
@@
 
+\bibitem{m1}
+	Marcuse, H. (2013). One-dimensional man: Studies in the ideology of advanced industrial society. Routledge.
+
+\bibitem{m2}
+	Marcuse, H. (2009). A responsabilidade da ciência. Scientiae studia, 7(1), 159-164.
+
+\bibitem{m3}
+	Marcuse, H. (2004). Technology, War and Fascism: Collected Papers of Herbert Marcuse (Vol. 1). Routledge.
+
+\bibitem{m4}
+	Pisani, M. M. (2009). Algumas considerações sobre ciência e política no pensamento de Herbert Marcuse. Scientiae Studia, 7(1), 135-158.
 \bibitem{au}
@@
 
+\bibitem{drones}
+	Pisani. M.
+	Entre Drones e Ciborgues: algumas considerações sobre teoria crítica e a necroética das redes sociais. Revista Impulso, Nov/2017.
+
+\bibitem{nexos1}
+	Pisani. M.
+	Entre Drones e Ciborgues: algumas considerações sobre teoria crítica e a necroética das redes sociais. Revista Impulso, Nov/2017.
+
+\bibitem{nexos2}
+	Dossies of the Nexos Network for Critical Theory and Interdisciplinary Research.
+Available at: \url{https://www.metodista.br/revistas/revistas-unimep/index.php/impulso/issue/view/207} and \url{https://www.metodista.br/revistas/revistas-unimep/index.php/impulso/issue/view/207}
+
 \bibitem{an}

### 2017-11-13 latex/nuvem/nuvem2.tex
meio revisao

@@
 \usepackage{nuvem}
+\usepackage[usenames,dvipsnames]{xcolor}
+\usepackage{hyperref}
+\hypersetup{
+	%pagebackref=true,
+	pdfcreator={LaTeX with abnTeX2},
+	pdfkeywords={abnt}{latex}{abntex}{USPSC}{trabalho acadêmico}, 
+	colorlinks=true,       		% false: boxed links; true: colored links
+	linkcolor=blue,          	% color of internal links
+	citecolor=blue,        		% color of links to bibliography
+	filecolor=magenta,      		% color of file links
+	urlcolor=blue,
+	allbordercolors=black,
+	bookmarksdepth=4
+}
+\usepackage[utf8]{inputenc}
 
@@
 	management, scientific and individual endeavors.
-	If you have data as information in a computer, and wants to
-	transmit it to an observer or user, the use of visual cues is an obvious,
-	and sometimes considered optimized, choice because of the efficiency and
+	If there is data, information or knowledge in a computer,
+	the observer or user should have it, the use of visual cues is an obvious,
+	and sometimes considered optimized, strategy because of the efficiency and
 	complexity of our visual cognition.
@@
 	due to patterns of music and spoken language.
-	How to correctly formalize the mapping of data to physical stimuli
+	How to correctly formalize the mapping of data, information
+	and knowledge to physical stimuli
 	and then to the perceived stimuli
-	respecting the Webner-Fechner and Steven's laws?
-	Related to sound, most traditional notation and musical qualities
-	respect the Webner-Fechner law, with note grids (e.g. semitones, octaves),
-	respecting the logarithm of frequency: iterval$=log_2(f_1/f_0)$,
-	loudness$=log_10(p_1/p_0)$
-	Duration seems to be conceptualized as such
-	because musical notation uses the division in two (simple, imperfectus),
-	amd three (perfectus) yielding a grid having
-	$d_i=d_{i+1}/2$ and $d_i=d_{i+1}/3$.
-	$durations=2^{x-i}$ and $=3^{x-i}$.
-	Which implies $log_2(dur) = x-i$.
-	A Webner-Fechner compliant relation
-	between the number of the durations and
-	depth of the division x-i = (4, 8, 16) for simple,
-	(3, 9, 12) for compound,
-	(5,7,13) complex.
-	6 is compound in traditional music theory,
-	but is the only one that relates periods of 2 and 3,
-	subdivision of 2 and 3.
+	respecting the Weber-Fechner and Steven's laws at least for visual
+	and auditory cues?
 \end{abstract}
@@
 The mapping of data to audiovisual cues
-involve a number of routines, conceptual frameworks,
+involves a number of routines, conceptual frameworks,
 and data.
@@
 poses ethical and conceptual challenges~\cite{an,an2}.
-An overview of the Linked Open Social Data gathered
-in RDF and related ontologies and vocabularies
+An overview of the Linked Open Social Data translated to
+RDF and related ontologies and vocabularies
 can be found in~\cite{nuvem1,tese,losd}.
@@
 \section{Current Framework}\label{current}
-We have strived to obtain relevant algebraic, empiric, conceptual,
+We obtained compelling algebraic, empiric and conceptual
 analyses of social structures~\cite{tese,stab},
-often supported by software, which is a very formal representation
-of what is being performed (although also very incomplete).
+often supported by software, which is a very efficient
+way to represent
+what is being performed (although also very incomplete).
 In this section, we focus on current efforts around Audiovisual Analytics.
 
-\subsection{AAVO}
+\subsection{AAVO}\label{sec:aavo}
 AAVO is a first tentative ontology for Audiovisual Analytics.
-Figure~\cite{aavo} depicts the AAVO Core,
-and both Core and Extended AAVO are well described in~\cite{aavoA}.
-The following concepts are envisioned for the core:
-
-And it might receive other layers of conceptualizations,
+Figure~\ref{aavo} depicts the AAVO Core,
+and both Core and Extended AAVO are well described in~\cite{aavo}.
+The following concepts are envisioned for the core
+but not yet implemented:
+Hypothesis, Analysis, Task/Purpose/Application.
+Some relations seem less fundamental then others,
+which might be expressed using the techniques described
+in~\cite{ont}.
+AAVO is very incipient, and will most probably receive
+other layers of conceptualizations beyond Core and Extended,
 such as for data type, colors, charts, etc.
 
+\begin{figure}[htbp]
+  \centering
+  \includegraphics[width=15cm]{../figs/aavomin}
+\caption{A graphical representation of current AAVO Core.
+	More information in Section~\ref{sec:aavo}.}\label{aavo}
+\end{figure}
+
 Its purposes are: automated inferences and recommendations on
-contexts involving data visualization;
+contexts involving data visualization (e.g. for suitable visualizations given a
+dataset or vice-versa);
 enable relevant theoretical discussions by having an
@@
 
-\subsection{An audiovisual analytics platform of social systems}
-A software system, a platform for audiovisual analytics of big data
-have been envisioned which have the following characteristics:
+\subsection{An audiovisual analytics platform for social systems}\label{sys}
+A software system, a web platform for audiovisual analytics of big data,
+has been envisioned with the following characteristics:
 \begin{itemize}
@@
 		of the audiovisualzations.
-		User might pause and set changing patterns..
-  \item The simultaneous use of both the vision and hearing channels.
+		User might pause and set changing patterns.
+  \item The simultaneous use of both the vision and auditory channels.
 	  User might mute or focus on the sonic cues.
@@
 	  suitable for scripting, one line commands, keyboard shortcuts.
+		It will probably be JavaScript with Vim-like one-liners and scripting.
 		(The written interface expresses a very broad space
@@
 		and textual data.
-  \item Designed to be enhanced with use by means of tests with users, of analyses of sessions and states
+  \item Designed to be enhanced with use by means of tests with users,
+	  and of analyses of sessions and states
 	  (maybe have facilities to record mouse movement, clicks and keyboard strokes).
@@
 
-The temporal dynamic of the interface is useful to do a rapid scan of the data by parts
+The temporal dynamic of the interface is useful to do a rapid scan of the data by parts.
 Observing data by parts is a core procedure in mining big data.
@@
 The focus on social systems entails analysis results of interest
-to both the scientific research, and State and private sectors.
-The artistic course of current and envisioned developments
-is useful for formal documents but also favors the engagement
+to scientific research, State and private sectors.
+Art in current and envisioned developments
+is useful for formal documents and favors the engagement
 of layman.
@@
 \subsection{Perceptual framework}
-Link to AAVO: the three sampling groups.
+Frequency and intensity are related to pitch and volume
+through the Weber-Fechner laws.
+One might also think about durations on the same sense.
+Visual estimuli, on the other hand, seem to be 
+more often modeled as the Steven's law.
+Is it possible to relate frequency and intensity to
+pitch and loudness through Steven's laws?
+Is the perception of visual cues also fit by the Weber-Fechner?
+The Steven's law is a power law while Weber-Fechner is an exponential
+law.
+This entails:
+\begin{equation}
+	\Delta p = log_X(e_1/e_0)
+\end{equation}
+\noindent for Weber-Fechner and
+\begin{equation}
+	p = k.e_1^\alpha
+\end{equation}
+\noindent for Steven's,
+where p is the (subjective) magnitude of the perceived stimuli,
+$X$ and $\alpha$ depend on the stimuli,
+and $e_0$ and $k$ depend on the the units used.
+
+In equally spaced samples of the stimuli (e.g. in PCM audio),
+the sample values might express amplitude or frequency.
+If the sample values represent durations, the separation
+between samples looses its meaning, but enables a Fourier
+analysis in a 'virtual' or 'imaginary' space.
+
+The possible equations relating physical and perceived stimuli
+should be integrated to AAVO.
+
+========
+
+	Some perception measurements are very consistently represented by
+	exponentil laws, such as betweeen pich and frequency.
+	Loudness is also more or less consistently related by exponentials
+	to the power, intensity or amplitude for a fixed frequency or spectrum.
+	Visual cues, on the other hand, present consistent power laws.
+
+	mention the human processor model.
+
+	Related to sound, most traditional notation and musical qualities
+	respect the Weber-Fechner law, with note grids (e.g. semitones, octaves),
+	respecting the logarithm of frequency: interval$=log_2(f_1/f_0)$,
+	loudness$=log_10(p_1/p_0)$.
+	Duration seems also to be conceptualized as such
+	because musical notation uses the division in two (simple, imperfectus),
+	amd three (perfectus) yielding a grid having
+	$d_i=d_{i+1}/2$ and $d_i=d_{i+1}/3$.
+	$durations=2^{x-i}$ and $=3^{x-i}$.
+	Which implies $log_2(dur) = x-i$.
+	A Weber-Fechner compliant relation
+	between the number of the durations and
+	depth of the division x-i = (4, 8, 16) for simple,
+	(3, 9, 12) for compound,
+	(5,7,13) complex.
+	6 is compound in traditional music theory,
+	but is the only one that relates periods of 2 and 3,
+	subdivision of 2 and 3.
+
 \subsection{Other: Erdös Sectioning, KS-derived statistics, Wordnet,
 text mining, stability and differentiation, partnerships}
+Other works, such as classification of participants (vertices) as
+hubs, intermediary and periphery, statistical tests e.g. to
+observe linguistic differences among the network participants,
+the use of Wordnet and token-oriented methods for text mining,
+and current partnership with researchers in physics, computer science,
+mathematics, philosopy and social sciences, clinical and social psychology,
+give some of the support for our endeavors.
+
 \subsection{What to do?}
@@
 
-\bibitem{krishnan00} E. Krishnan, A. M. Shan, T. Rishi, L. A. Ajith, C. V.
-Radhakrishnan, \textit{On-line Tutorial on \LaTeX{}},
-``Mathematics'' (Indian \TeX{} Users Group, 2000), \\
-\url{http://www.tug.org/tutorials/tugindia/chap11-scr.pdf}.
+\bibitem{an}
+	Fabbri, R. What are you and I? [anthropological physics fundamentals], 2015. Available at \url{https://www.academia.edu/10356773/What\_are\_you\_and\_I\_anthropological\_physics\_fundamentals\_}
 
-\bibitem{vantrigt97} C. van Trigt, ``Visual system-response functions and estimating reflectance,''
-J. Opt. Soc. Am. A \textbf{14}, 741--755 (1997).
+\bibitem{an2}
+	Anthropological physics and social psychology in the critical research of networks. Complex Networks Digital Campus (CS-DC'15).
+	Em \\\url{http://cs-dc-15.org/papers/cognition/social-psychology/anthropological-physics-and-social-psychology-in-the-critical-research-of-networks/}
 
-\bibitem{masters93} T. Masters, \emph{Practical Neural Network Recipes in C++} (Academic, 1993).
+\bibitem{nuvem1} 
+  R.~Fabbri and M. Pisani
+  ``Analytic consideration of the society by itself,''
+  I Workshop @NUVEM, UFABC campus de Santo André,
+		Available at: \url{https://github.com/ttm/aavo/raw/master/latex/nuvem/nuvem1.pdf}.
 
-\bibitem{shoop97} B. L. Shoop, A. H. Sayles, and D. M. Litynski, ``New devices for optoelectronics: smart pixels,''
-in \emph{Handbook of Fiber Optic Data Communications},
-C. DeCusatis, D. Clement, E. Maass, and R. Lasky, eds. (Academic, 1997), pp. 705--758.
+\bibitem{tese}
+Fabbri, R. (2017). Topological stability and textual differentiation in human interaction networks:
+		statistical analysis, visualization and linked data. Doctoral thesis.
+		From \url{https://github.com/ttm/thesis/raw/master/thesis-rfabbri.pdf}
 
-\bibitem{kalman76} R. E. Kalman,``Algebraic aspects of the generalized inverse of a rectangular matrix,'' in
-\emph{Proceedings of Advanced Seminar on Generalized Inverse and Applications}, M. Z. Nashed, ed. (Academic, 1976), pp. 111--124.
+\bibitem{losd}
+	Fabbri, R., \& de Oliveira, O. N. (2016). Linked Open Social Database. Github repositories, from \url{https://github.com/ttm/linkedOpenSocialDatabase/raw/master/paper.pdf}
 
-\bibitem{craig96} R. Craig and B. Gignac, ``High-power 980-nm pump lasers,''
-in \emph{Optical Fiber Communication Conference}, Vol. 2 of 1996 OSA Technical Digest Series (Optical Society of America, 1996), paper ThG1.
+\bibitem{ont} 
+  R.~Fabbri and M. Pisani,
+  ``Enhancements of linked data expressiveness for ontologies,''
+  Anais do XX ENMC - Encontro Nacional de Modelagem Computacional e
+  VIII ECTM - Encontro de Ci\^encias e Tecnologia de Materiais, Nova Friburgo,
+  RJ - 16 a 19 Outubro 2017
+		Available at: \url{https://github.com/ttm/equanimousScalefree/raw/master/article/article.pdf}.
 
-\bibitem{steup96} D. Steup and J. Weinzierl, ``Resonant THz-meshes,''
-presented at the Fourth International Workshop on THz Electronics, Erlangen-Tennenlohe, Germany, 5--6 Sept. 1996.
+\bibitem{stab}
+Fabbri, R., Fabbri, R., Antunes, D. C., Pisani, M. M., \& de Oliveira, O. N. (2017). Temporal stability in human interaction networks. Physica A: Statistical Mechanics and its Applications.
+
+\bibitem{aavo} 
+  R.~Fabbri,
+		``Audiovisual Analytics Vocabulary and Ontology (AAVO): initial core and example expansion ,''
+  Anais do XX ENMC - Encontro Nacional de Modelagem Computacional e
+  VIII ECTM - Encontro de Ci\^encias e Tecnologia de Materiais, Nova Friburgo,
+  RJ - 16 a 19 Outubro 2017
+		Available at: \url{https://arxiv.org/abs/1710.09954}.
+
+\bibitem{tufte}
+	Tufte, E. R. (1990). Envisioning information. Graphics press.
 

### 2017-11-13 latex/nuvem/nuvem2.tex
finalizada primeira revisao

@@
 Frequency and intensity are related to pitch and volume
-through the Weber-Fechner laws.
+through the Weber-Fechner law.
 One might also think about durations on the same sense.
 Visual estimuli, on the other hand, seem to be 
-more often modeled as the Steven's law.
+more often modeled as the Steven's law (is this true?).
 Is it possible to relate frequency and intensity to
-pitch and loudness through Steven's laws?
-Is the perception of visual cues also fit by the Weber-Fechner?
+pitch and loudness through Steven's law?
+Is the perception of visual cues also fit by Weber-Fechner?
 The Steven's law is a power law while Weber-Fechner is an exponential
-law.
-This entails:
+law:
 \begin{equation}
@@
 and $e_0$ and $k$ depend on the the units used.
+$\Delta p$ is an interval, a distance, in the perceived magnitude.
 
@@
 analysis in a 'virtual' or 'imaginary' space.
+For example, consider the Fouriter transform of
+this sequence of durations in seconds:
+[1, 0.5, 0,5, 1, 0.5, 0.5].
+The components are sinusoids whose amplitude
+is duration.
+The durations are the sample values,
+each value corresponding to a sample separated
+by a virtual distance.
+% keep static amplitude or phase
+% in virtual and real-simed Fourier transfoms
 
-The possible equations relating physical and perceived stimuli
-should be integrated to AAVO.
-
-========
-
-	Some perception measurements are very consistently represented by
-	exponentil laws, such as betweeen pich and frequency.
-	Loudness is also more or less consistently related by exponentials
-	to the power, intensity or amplitude for a fixed frequency or spectrum.
-	Visual cues, on the other hand, present consistent power laws.
-
-	mention the human processor model.
-
-	Related to sound, most traditional notation and musical qualities
+Related to sound, most traditional music notation and qualities
 respect the Weber-Fechner law, with note grids (e.g. semitones, octaves),
@@
 $durations=2^{x-i}$ and $=3^{x-i}$.
-	Which implies $log_2(dur) = x-i$.
+Which implies $log_2(dur) = x-i \Rightarrow i = log_2(X/dur), x = log_2X$.
 A Weber-Fechner compliant relation
@@
 but is the only one that relates periods of 2 and 3,
-	subdivision of 2 and 3.
+
+The Human Processor Model~\cite{HPM}
+has somewhat established values for
+a number of cognitive characteristics such as eye movement time (230ms),
+half-life of visual image storage (200ms) and 
+of
+working memory (7s),
+and visual and auditory capacities (17 letters, 5 letters).
+These reference values, together with the
+equations relating physical and perceived stimuli,
+should be integrated to AAVO.
 
@@
 text mining, stability and differentiation, partnerships}
-Other works, such as classification of participants (vertices) as
-hubs, intermediary and periphery, statistical tests e.g. to
-observe linguistic differences among the network participants,
-the use of Wordnet and token-oriented methods for text mining,
+Other works, such as the classification of participants (vertices) as
+hubs, intermediary and peripheral; statistical tests e.g. to
+observe linguistic differences among the network participants;
+the use of Wordnet and token-oriented methods for text mining;
 and current partnership with researchers in physics, computer science,
@@
 \subsection{What to do?}
-We will surely integrate the Hypothesis, Goals, Interaction
-concepts into the AAVO Core.
+We will surely integrate the envisioned concepts into AAVO Core.
 We should develop the system in Section~\ref{sys} to some extent,
@@
 With the research groups @Nuvem, 
+especially those audiovisual and societal,
 relate the missing concepts in AAVO,
-enhance the software design of the system,
+enhance the software design of the system.
 Know about the right tools for visualizing
@@
 and known conceptualizations.
-Virtual reality, visualization and societal groups
 might want to formalize their own conceptualizations,
@@
 or which are confluent (e.g. symbiotic)
-with Section~\ref{current}: software
+with Section~\ref{current}: software,
 theoretical framework, research and institutional experiments and equipment,
@@
 
+\bibitem{HPM}
+	Human processor model.
+(2017, March 11).
+In Wikipedia, The Free Encyclopedia.
+Retrieved 06:16, November 13, 2017, from \url{https://en.wikipedia.org/w/index.php?title=Human_processor_model&oldid=769706845}
+
 \end{thebibliography}

### 2017-11-13 latex/nuvem/nuvem1.tex
meio revisao
feita revisao do nuvem1

@@
 \begin{abstract}
-Social participation has been growing on the last decade though with a considerable
-drawback with the recent rise of conservative political parties 
+Social participation has been growing on the last decade despite
+the recent rise of conservative political parties 
 in various locations of the world.
@@
 And the consideration of the authoritarian personality traces~\cite{au}
-have gained relevance due to unforeseen social phenomena
+has gained relevance due to unforeseen social phenomena
 such as the recent and worldwide rise of the right-winged manifestations
@@
 There are many ways to achieve this,
-and it is performed to some extent~\cite{scriptLattes,losd,p5,an}.
+and it is performed to some extent~\cite{scriptLattes,losd,p5,an,an2}.
 Our focus here is on the mining of the data
@@
 Cidade Democrática and AA have been translated into linked data (RDF)
-and has been named the LOD (Linked Open Social Data~\cite{losd}).
+and has been named the LOSD (Linked Open Social Data~\cite{losd}).
 Initial OWL ontologies were derived from the data by automated routines.
@@
 Even so, it has been used e.g. by participants of a SESC workshop
-and has received some attention from the scientific community~\cite{mariliaNexos,tese},
+and has received some attention from the scientific community~\cite{drones,tese},
 and mining of the data revealed very stable patterns as predicted by natural laws
@@
  automated inference; and data organization, integration and linkage by means of conceptualizations.
-It seems reasonable to use the ontologies to link the social participation data
+It seems reasonable to use these ontologies to link the social participation data
 and enable a scalable participatory legacy of machine- and human- friendly representation
-of data for navigation, query and analysis, automated inference
-and conceptual considerations.
+of data.
 
@@
 were considered in~\cite{imp,stab,tese}.
-In summary, the hub, intermediary and peripheral sectorialization
-of a network might be done in a social network.
+In summary, the hubs, intermediary and periphery sectorialization
+of a network might be achieved (or imposed) in a social network.
 It yields a classification of the participants
@@
 of a quantitative classification (or typology)
-of human beings..
-At the same time, the roles participant
+of human beings.
+At the same time, the classification of the participants
+(hub, intermediary or peripheral)
 vary in time constantly~\cite{stab,tese}
@@
 phenomena in the social structures
-of the observer him/herself and has been
-considered in.
+of the observer and has been
+considered in~\cite{an,an2,tese}.
 Comte talked about anthropological physics,
@@
 It is based on voluntary logging of messages (called shouts)
-about what is being done (reading, coding, etc) that might
+about what is being done (reading, coding, etc).
+The Shouts may
 be linked to sessions and screencasts or be blind-reviewed
@@
 traditional documents (e.g. articles, books) and participation (e.g. conferences, thesis defense),
-and currently performed e.g. by Lattes.
+as currently performed e.g. by Lattes Curriculum.
 Such documentation (log of dedications)
@@
 extreme poverty, as the poor would be paid
-to act and change State measures for the poor for
+to act and change State measures for the poor
 or a community.
@@
 The developments for Audiovisual Analytics are incipient
-and has the development focus of only one researcher and developer.
+and have the development focus of only one researcher and developer.
 The Nexos research network is very active nation wide.
@@
 and potential enhancements through criticism of its organization,
-which is valuable for the technological developments being made.
+which is valuable for the technological developments being made~\cite{nexos1}.
 
@@
 in Brazil? Ontologies?
-We might use directions 
+We might benefit from directions 
 on better linking LOSD to the semantic web
@@
 software
-(persistence, web?, analysis methods, audiovisual rendering).
+(persistence, web?, analysis methods, audiovisual rendering, etc,
+as described in~\cite{nuvem2}).
 
-Other questions we have that might benefit from discussion and entail collaboration are:
+Other questions that might initiate
+nice discussions or entail collaboration are:
 how to enable a self-transparency user base?
@@
 that enables the user base and interested parties to
-collective gather social data, analysis and conceptualizations?
+collective gather social data, analyses and conceptualizations?
 Is the societal consideration of our social data
@@
 Is it possible to achieve social participation in the private sector,
-e.g. to have civil representation in the chair of companies,
+e.g. to have civil representation in the management of companies,
 to regulate matters such as YouTube adds in Google,
@@
 
+\bibitem{p5}
+Fabbri, R. Social participation ontologies and rules to fuel a social participation linked data cloud, 2014.
+	Technical report for the United Nations Development Program, made with the Brazilian Presidency.
+		Available at \url{https://github.com/ttm/pnud5/raw/master/latex/produto.pdf}
+
+\bibitem{an}
+	Fabbri, R. What are you and I? [anthropological physics fundamentals], 2015. Available at \url{https://www.academia.edu/10356773/What\_are\_you\_and\_I\_anthropological\_physics\_fundamentals\_}
+
+\bibitem{an2}
+	Anthropological physics and social psychology in the critical research of networks. Complex Networks Digital Campus (CS-DC'15).
+	Em \url{http://cs-dc-15.org/papers/cognition/social-psychology/anthropological-physics-and-social-psychology-in-the-critical-research-of-networks/}
+
 \bibitem{kolm} 
@@
 
-\bibitem{p5}
-Fabbri, R. Social participation ontologies and rules to fuel a social participation linked data cloud, 2014.
-	Technical report for the United Nations Development Program, made with the Brazilian Presidency.
-		Available at \url{https://github.com/ttm/pnud5/raw/master/latex/produto.pdf}
-
-\bibitem{p5}
-Fabbri, R. Social participation ontologies and rules to fuel a social participation linked data cloud, 2014.
-	Technical report for the United Nations Development Program, made with the Brazilian Presidency.
-		Available at \url{https://github.com/ttm/pnud5/raw/master/latex/produto.pdf}
+\bibitem{drones}
+	Pisani. M.
+	Entre Drones e Ciborgues: algumas considerações sobre teoria crítica e a necroética das redes sociais. Revista Impulso, Nov/2017.
 
@@
 
+\bibitem{ont} 
+  R.~Fabbri and M. Pisani,
+  ``Enhancements of linked data expressiveness for ontologies,''
+  Anais do XX ENMC - Encontro Nacional de Modelagem Computacional e
+  VIII ECTM - Encontro de Ci\^encias e Tecnologia de Materiais, Nova Friburgo,
+  RJ - 16 a 19 Outubro 2017
+		Available at: \url{https://github.com/ttm/equanimousScalefree/raw/master/article/article.pdf}.
+
 \bibitem{imp}
@@
 		\url{https://www.metodista.br/revistas/revistas-unimep/index.php/impulso/article/view/3367}.
-
-\bibitem{ont} 
-  R.~Fabbri and M. Pisani,
-  ``Enhancements of linked data expressiveness for ontologies,''
-  Anais do XX ENMC - Encontro Nacional de Modelagem Computacional e
-  VIII ECTM - Encontro de Ci\^encias e Tecnologia de Materiais, Nova Friburgo,
-  RJ - 16 a 19 Outubro 2017
-		Available at: \url{https://github.com/ttm/equanimousScalefree/raw/master/article/article.pdf}.
 
@@
 
-\bibitem{drones}
-	Pisani. M.
-	Entre Drones e Ciborgues: algumas considerações sobre teoria crítica e a necroética das redes sociais. Revista Impulso, Nov/2017.
-
 \bibitem{nexos1}
-	Pisani. M.
-	Entre Drones e Ciborgues: algumas considerações sobre teoria crítica e a necroética das redes sociais. Revista Impulso, Nov/2017.
-
-\bibitem{nexos2}
 	Dossies of the Nexos Network for Critical Theory and Interdisciplinary Research.
 Available at: \url{https://www.metodista.br/revistas/revistas-unimep/index.php/impulso/issue/view/207} and \url{https://www.metodista.br/revistas/revistas-unimep/index.php/impulso/issue/view/207}
-
-\bibitem{an}
-	Fabbri, R. What are you and I? [anthropological physics fundamentals], 2015. Available at \url{https://www.academia.edu/10356773/What\_are\_you\_and\_I\_anthropological\_physics\_fundamentals\_}
-
-\bibitem{an2}
-	Anthropological physics and social psychology in the critical research of networks. Complex Networks Digital Campus (CS-DC'15).
-	Em \url{http://cs-dc-15.org/papers/cognition/social-psychology/anthropological-physics-and-social-psychology-in-the-critical-research-of-networks/}
 
@@
   I Workshop @NUVEM, UFABC campus de Santo André,
-		Available at: \url{}.
+		Available at: \url{https://github.com/ttm/aavo/raw/master/latex/nuvem/nuvem2.pdf}.
 \end{thebibliography}

### 2017-11-13 latex/nuvem/nuvem1.tex
spell

@@
 (although the intermediary are also reported as authorities
-and structuraly most important agents)~\cite{ega},
+and structurally most important agents)~\cite{ega},
 which favors the potential for prejudice
@@
 into performing the social participation.
-The existence of such a cycle would erradicate
+The existence of such a cycle would eradicate
 extreme poverty, as the poor would be paid

### 2017-11-13 latex/nuvem/nuvem2.tex
spell

@@
 	If there is data, information or knowledge in a computer,
-	the observer or user should have it, the use of visual cues is an obvious,
+	and the observer or user should have it, the employment of visual cues is an obvious,
 	and sometimes considered optimized, strategy because of the efficiency and

### 2017-11-13 latex/nuvem/nuvem2.tex
terminado de revisar o nuvem2

@@
 	complexity of our visual cognition.
-	Our sonic cognition is sometimes considered of greater complexity,
+	Our auditory cognition is sometimes considered of greater complexity,
 	due to patterns of music and spoken language.
@@
 	and auditory cues?
+	How to correctly express data visualization theory and practice,
+	relating datatypes to processing routines and suitable audiovisualizations?
 \end{abstract}
@@
 and data.
-The scaling and integration factors are decisive in setting
-of Big Data such as in social data mining.
+The scaling and integration factors are decisive in settings
+of big data such as in social data mining.
 The study of data that expresses human social systems
@@
 and both Core and Extended AAVO are well described in~\cite{aavo}.
-The following concepts are envisioned for the core
+The following concepts/classes are envisioned for the core
 but not yet implemented:
@@
 		of possible commands.)
-  \item Linked data to achieve a formal and browse-friendly representation of data for both humans and machines;
+  \item Linked data to achieve a formal and browsing-friendly representation of data for both humans and machines;
 	  for a unified consideration of both data and knowledge architecture;
@@
 
-The temporal dynamic of the interface is useful to do a rapid scan of the data by parts.
+The temporal dynamic of the interface is useful to make a rapid scan of the data by parts.
 Observing data by parts is a core procedure in mining big data.
@@
 Complex networks and text yield a broad user base
-and set of knowledge fields which might be studied withing the system.
-The focus on social systems entails analysis results of interest
+and set of knowledge fields which might be studied with help of such system.
+The focus on social structures entails analysis results of interest
 to scientific research, State and private sectors.
 Art in current and envisioned developments
-is useful for formal documents and favors the engagement
+is useful for favoring an attentive interaction of the user,
+the enhancement of formal documents and the engagement
 of layman.
@@
 One might also think about durations on the same sense.
-Visual estimuli, on the other hand, seem to be 
-more often modeled as the Steven's law (is this true?).
+Visual stimuli, on the other hand, seem to be 
+more often modeled by the Steven's law (is this true?).
 Is it possible to relate frequency and intensity to
@@
 $\Delta p$ is an interval, a distance, in the perceived magnitude.
+$x$, $\alpha$, $e_0$ and $k$ are usually considered constants.
 
@@
 each value corresponding to a sample separated
-by a virtual distance.
+by a virtual (unitary) distance.
 % keep static amplitude or phase
@@
 respecting the logarithm of frequency: interval$=log_2(f_1/f_0)$,
-loudness$=log_10(p_1/p_0)$.
+loudness$=log_{10}(p_1/p_0)$.
 Duration seems also to be conceptualized as such
@@
 $durations=2^{x-i}$ and $=3^{x-i}$.
-Which implies $log_2(dur) = x-i \Rightarrow i = log_2(X/dur), x = log_2X$.
+Which implies $log_2(dur) = x-i \Rightarrow i = log_2(X/dur),\;x = log_2X$.
 A Weber-Fechner compliant relation
@@
 mathematics, philosopy and social sciences, clinical and social psychology,
-give some of the support for our endeavors.
+give some of the support for our endeavors.~\cite{tese}
 
@@
 especially those audiovisual and societal,
-relate the missing concepts in AAVO,
-enhance the software design of the system.
-Know about the right tools for visualizing
+we might relate the missing concepts in AAVO,
+or enhance the software design of the system;
+try to find more about the right tools for visualizing
 ontologies and social data.
-For knowing about pertinent analysis methods
+about pertinent analysis methods
 and known conceptualizations.
-might want to formalize their own conceptualizations,
+@Nuvem researchers might want to formalize their own conceptualizations,
 and concepts related to AAVO might have similar developments
-or which are confluent (e.g. symbiotic)
+or be confluent (e.g. symbiotic)
 with Section~\ref{current}: software,

### 2017-11-14 latex/nuvem/nuvem1.tex
pequena correcao em nuvem1

@@
 Comte talked about anthropological physics,
-but it had the sense which physical (or biological or natural)
+but it had the sense which physical
 anthropology has, which is a branch of anthropology
-(the other is social anthropology).
+(the other is sociocultural anthropology).
 Anthropological physics, in the sense we use and understand correct,

### 2017-11-20 latex/nuvem/nuvem1.tex
retoques no nuvem1

@@
 \title{Analytic consideration of the society by itself}
+ANALYTIC CONSIDERATION OF THE SOCIETY BY ITSELF
 
@@
 \begin{abstract}
-Social participation has been growing on the last decade despite
-the recent rise of conservative political parties 
-in various locations of the world.
 We aim to consider a set of developments on linked social data
-and analysis performed over the last years, with a focus
+and analyses performed over the last years, with a focus
 on the potential use for the civil society and scientific academy.
@@
 
+\subsection{Acknowledgments}
+FAPESP (project 2017/05838-3); ICMC/USP and UFABC researchers;
+the open source and free software communities that
+made these developments possible.
+
 \begin{thebibliography}{99}

### 2017-11-20 latex/nuvem/nuvem2.tex
retoques no nuvem1

@@
 \author{Renato Fabbri$^1$\\
-	Marilia M. Pisani$^2$
+	Marilia M. Pisani$^2$\\
+	Maria Cristina Ferreira de Oliveira$^3$
 }
-\address{$^1$IFSC/USP, $^2$CCNH/UFABC}
-\email{$^1$renato.fabbri@gmail.com, $^2$marilia.m.pisani@gmail.com}
+\address{$^{1,3}$IFSC/USP, $^2$CCNH/UFABC}
+\email{$^1$renato.fabbri@gmail.com, $^2$marilia.m.pisani@gmail.com,
+$^3$cristina@icmc.usp.br}
 
@@
 
-\subsection{Perceptual framework}
+\subsection{Perceptual framework}\label{perceptual}
 Frequency and intensity are related to pitch and volume
@@
 \subsection{Other: Erdös Sectioning, KS-derived statistics, Wordnet,
-text mining, stability and differentiation, partnerships}
+text mining, stability and differentiation, partnerships}\label{other}
 Other works, such as the classification of participants (vertices) as
@@
 With the research groups @Nuvem, 
-especially those audiovisual and societal,
+especially with those more related to audiovisual and social issues,
 we might relate the missing concepts in AAVO,
@@
 try to find more about the right tools for visualizing
-ontologies and social data.
-about pertinent analysis methods
-and known conceptualizations.
+ontologies and social data, or about
+pertinent analysis methods and known conceptualizations.
 @Nuvem researchers might want to formalize their own conceptualizations,
@@
 linked data, and other initiatives or knowledge.
+Help on correctly describing the mapping of data to physical
+and cognitive stimuli, as in Section~\ref{perceptual},
+will be highly appreciated.
+We might share and consider other supporting resources of our
+research, e.g. those in Section~\ref{other}, upon
+interest.
+
+\subsection{Acknowledgments}
+FAPESP (project 2017/05838-3); ICMC/USP and UFABC researchers;
+the open source and free software communities that
+made these developments possible.
 
